Vaccination with mycobacterial lipid loaded nanoparticle leads to lipid antigen persistence and memory differentiation of antigen-specific T cells

Version 1

Abstract

Abstract

Mycobacterium tuberculosis (Mtb) infection elicits both protein and lipid antigen-specific T cell responses. However, the incorporation of lipid antigens into subunit vaccine strategies and formulations has been under-explored, and the properties of vaccine-induced Mtb lipid-specific memory T cells have remained elusive. Mycolic acid (MA), a major lipid component of the Mtb cell wall, is presented by human CD1b molecules to unconventional T cell subsets. These MA-specific CD1b-restricted T cells have been detected in the blood and disease sites of Mtb-infected individuals, suggesting that MA is a promising lipid antigen for incorporation into multicomponent subunit vaccines. In this study, we utilized the enhanced stability of bicontinuous nanospheres (BCN) to efficiently encapsulate MA for delivery in vivo to MA-specific T cells both alone and in combination with an immunodominant Mtb protein antigen (Ag85B). Pulmonary delivery of MA-loaded BCN (MA-BCN) elicited MA-specific T cell responses in humanized CD1 transgenic mice. Simultaneous delivery of MA and Ag85B within BCN activated both MA- and Ag85B-specific T cells. Interestingly, pulmonary vaccination with MA-Ag85B-BCN led to the persistence of MA, but not Ag85B, within alveolar macrophages in the lung. Vaccination of MA-BCN through intravenous or subcutaneous route, or with attenuated Mtb likewise reproduced MA persistence. Moreover, MA-specific T cells in MA-BCN-vaccinated mice differentiated into a T follicular helper-like phenotype. Overall, the BCN platform allows for the dual encapsulation and in vivo activation of lipid and protein antigen-specific T cells and leads to persistent lipid depots that could offer long-lasting immune responses.

Introduction

Vaccination is currently the most effective method for the prevention and eradication of infectious diseases. These efforts have been enhanced by the development of subunit vaccine formulations that include only immunodominant antigens from pathogens paired with select adjuvants. In contrast to attenuated and inactivated vaccines that contain diverse molecular components of pathogens, subunit vaccines are often preferred due to their simplicity, safety, stability, and scalability of production [1]. Yet the development of effective vaccines for several key endemic parasitic and bacterial pathogens has remained elusive. Antigen selection for subunit vaccines has thus far focused on protein molecules, overlooking a vast family of non-protein antigens such as lipids/glycolipids, metabolites, and phosphoantigens, which can be found on and within pathogens and can be recognized by various subsets of unconventional T cells [2].

Persistent antigen exposure is known to occur naturally after infection through antigen archiving by follicular dendritic cells (FDCs), which periodically present antigen to cognate B cells and aid in affinity maturation and somatic hypermutation [3, 4]. Antigen persistence can also support the maintenance of T cell response through the periodic priming of circulating T cells [5–10]. Subcutaneous vaccination has also been shown to lead to archiving of protein antigens within lymphatic endothelial cells (LECs), which induce the development of protective T cell immunity [11, 12]. However, little is known about how other vaccination routes and vaccine formulations may affect antigen archiving and persistence.

Lipid and glycolipid antigens can be presented by group 1 (CD1a, b, and c) and group 2 (CD1d) CD1 molecules to CD1-restricted T cells. Unlike MHC molecules, CD1 molecules exhibit limited polymorphism [13]. Therefore, CD1-restricted microbial lipid antigens are likely to be recognized by most individuals, making them attractive targets for vaccine development [14]. Group 1 CD1-restricted T cells have diverse T cell receptors and can recognize a variety of lipid and glycolipid antigens [15]. Microbial lipid-specific group 1 CD1-restricted T cells have been identified from several pathogens, including Mtb and other mycobacterial species: Staphylococcus aureus (SA), Borrelia burgdorferi, Salmonella typhimurium, and Brucella melitensis [15]. CD1b-restricted T cells specific to Mtb lipids including mycolic acid (MA), an immunodominant lipid antigen[16] and component of the mycobacterial cell wall, have been identified in patients with tuberculosis (TB) [17, 18]. As group 1 CD1 molecules are not expressed in mice, we have previously developed a human CD1 transgenic mouse model (hCD1Tg). [19] Using hCD1Tg mice, we have shown that transgenic CD1b-restricted DN1 T cells specific for MA could protect against Mtb infection [20], making these potential antigens of interest in a subunit vaccine against TB.

The primary goal of a vaccine is to induce the differentiation of the adaptive immune system such that upon re-exposure to the antigen, a faster and larger response occurs, halting pathogen spread. While conventional T cells and B cells are known to undergo this differentiation, unconventional T cells such as NKT cells do not possess a comparable memory phenotype. The presence of Mtb lipid-specific group 1 CD1-restricted memory T cells has been described in both human and mice [17–19]. However, the properties of memory group 1 CD1-restricted T cells, which share similar developmental pathway as NKT cells, are not known.

Lipid antigens are difficult to formulate and administer due to their hydrophobic nature. To assess the utility of MA as a component of a subunit vaccine, we previously encapsulated MA in poly(ethylene glycol)-block-poly(propylene sulfide) (PEG-b-PPS) micellar nanocarriers (MA-MC) [21]. Nanocarriers composed of PEG-b-PPS possess several unique benefits compared to other delivery platforms, including high stability in aqueous solution [22], minimal background immunostimulation [23], and triggered release of cargo when exposed to physiological levels of reactive oxygen species [24–26]. We showed MA-MC could effectively activate adoptively transferred DN1 T cells and elicit polyclonal group 1 CD1-restricted T cell response in hCD1Tg mice after intranasal delivery [21]. While effective for stable loading of lipophilic and amphiphilic payloads, micellar nanocarriers, do not allow for facile encapsulation of hydrophilic payloads like protein antigens.

Here, we build upon our prior lipid antigen vaccine formulation by employing a nanocarrier platform engineered for the dual loading of both lipid and protein antigens. Bicontinuous nanospheres (BCN) are polymeric analogs to lipid cubosomes and thus possess a highly organized internal cubic organization containing both intertwined hydrophobic bilayers for lipid loading as well as hydrophilic aqueous channels for protein loading [27]. Using the flash nanoprecipitation (FNP) technique [28, 29], we fabricated MA-loaded BCN (MA-BCN) and found them to have superior stimulatory capacity in vivo compared to MA loaded poly(D,L-lactide-co-glycolide) (PLGA) nanocarriers (MA-PLGA), which is a widely used nanocarrier platform for Mtb vaccine development [30, 31]. Furthermore, we found that after MA-BCN vaccination, MA could persist for 6 weeks post-vaccination within alveolar macrophages, a phenomenon which was dependent on the presence of an encapsulating vector, nanocarrier or bacterial, but not the route of vaccination. Due to the enhanced stability of the BCN architecture that can support the loading of diverse and multiple payloads [22, 32], we were able to efficiently co-encapsulate within BCN both MA and Ag85B, an immunodominant protein antigen of Mtb [33]. Interestingly, while both antigens activated their corresponding antigen-specific T cells, only MA resulted in antigen persistence, suggesting a potentially unique characteristic of subunit vaccines containing lipid antigens.

Results

BCN morphology preserved after MA loading

We scalably assembled spherical BCN with and without loaded MA using flash nanoprecipitation (FNP) as previously described [29]. Dynamic light scattering (DLS) showed that both MA-BCN and BCN were consistent in size (364 ±19 nm and 354 ± 14 nm, respectively), and were monodisperse based on their polydispersity indices (PDIs) of 0.24 ± 0.04 and 0.21 ±0.05, respectively (Fig 1A). We next verified that the BCN formulation maintained its characteristic interconnected aqueous channels using cryogenic transmission electronic microscopy (cryo-TEM) (Fig 1B) and negative staining transmission electron microscopy (TEM) (Fig 1C). Using small angle X-ray scattering (SAXS) studies, we confirmed the internal cubic organization of BCN. Bragg peaks at the √2, √4, and √6 ratios show that the primitive type of cubic internal organization was preserved between MA-BCNs and BCNs (Fig 1D). Thus, MA encapsulation did not disturb the BCN architecture. We also manufactured a PLGA nanocarrier formulation (PLGA-NP) with and without MA encapsulation. PLGA-NP morphology were confirmed using cryoTEM (Suppl Fig 1A) and SAXS (Suppl Fig 1B) and the size of blank PLGA-NP and MA-loaded PLGA-NP (MA-PLGA) were found to be 169 ± 4 and 163 ± 5 nm, respectively, with PDIs of 0.18 ± 0.04 and 0.19 ± 0.12, respectively (Fig 1E). Using the coumarin derivatization method [21], we found that BCNs could more efficiently encapsulate MA than PLGA-NPs (Fig 1F). We previously noted that the highly stable cubic architecture of BCN can lead to the retention of cargo within the endolysosomal compartment [22]. Indeed, Texas Red-Dextran loaded BCN were mostly found in lysosomes while Texas Red-Dextran loaded PLGA-NP could be found both within the lysosome and cytosol of cells (Supp Fig 2A, B).

Physicochemical characterization of BCN and PLGA-NP.

(A) Dynamic light scattering (DLS) analysis of blank BCN and MA-BCN. (B) CryoTEM of MA-BCN (scale = 500 nm). (C) Negative staining TEM images of MA-BCN. (D) SAXS of blank BCN and MA-BCN. (E) DLS analysis of blank PLGA and MA PLGA. (F) MA encapsulation efficiency for MA-BCN and MA PLGA. N = 3 per condition. Data represented as mean ± SD.

MA-BCN activate MA-specific T cells in vitro and in vivo

As PLGA has served as a component of a wide range of vaccine formulations, including Mtb vaccines [30, 31], we benchmarked PEG-b-PPS BCN against PLGA-NP as a nanocarrier platform for lipid antigens. Both MA-BCN and MA-PLGA were highly effective in activating both mouse (DN1) and human (M11) MA-specific T cells in vitro and inducing IFN-γ production (Fig 2A-D). In particular, MA-BCN were significantly better at activating MA-specific T cells compared to free MA at equivalent concentrations (Fig 2A, B and D). In comparison to MA-BCN, MA-PLGA was significantly more effective at stimulating mouse MA-specific T cells in vitro at lower concentrations. Interestingly, even in the absence of MA, PLGA-NP showed strong stimulation particularly for M11 T cells. These results demonstrate that MA encapsulation within BCN enhances its ability to stimulate antigen-specific T cells and reveal a considerable background, and thus difficult to control, stimulatory effect of blank PLGA-NP. In contrast, blank BCN showed little background immunostimulation [23], allowing a more controlled dose-dependent increase in T cell responses as the loaded MA concentration increased.

MA-BCN effectively activated MA-specific T cells in vitro and in vivo.

(A, B) BMDCs were pulsed for 18 hrs. with selected nanoparticles at various concentrations, co-cultured for 48 hrs. with DN1 T cells, and T cell activation was measured. (A) Percentage of CD69 and CD25-expressing DN1 T cells. (B) IFN-g production measured by ELISA. (N=3). (C, D) ELISPOT of IFN-g from MA-specific human M11 T cells cultured for 16 hours with monocyte-derived dendritic cells with selected nanoparticles at high (C) and low (D) concentrations. Data are representative of two independent experiments and displayed as mean ± SEM. Statistical analysis: 2-way ANOVA, significance designated for MA-BCN vs free MA. (E-G) hCD1Tg mice were IN vaccinated with selected nanoparticles and CellTrace Violet-labeled DN1 T cells were adoptively transferred the next day. After 1 week, DN1 T cell activation and proliferation was measured. (E) Representative FACS plots of DN1 T cells in the LN. Percentage of proliferating (F) and CD44-expressing (G) DN1 T cells in the LN and lung. Data represented as mean ± SEM. Statistical analysis: 2-way ANOVA. *p<0.05, **p<0.01, ***p<0.001, ****p< 0.0001.

We next tested the ability of MA-loaded and unloaded BCN and PLGA to stimulate DN1 T cells in vivo. Intranasal (IN) vaccination of hCD1Tg mice was followed by the adoptive transfer of DN1 T cells. Surprisingly, we found that vaccination with MA-BCN induced a significantly higher percent of proliferation and activation in DN1 T cells than MA-PLGA in the draining lymph nodes (LN) and lungs at 1 week post-vaccination (Fig 2E-G). The extent of cell proliferation induced by MA-PLGA was not significantly different from that of blank PLGA (Fig 2F). Since MA-BCN effectively stimulated MA-specific T cells in vivo while MA-PLGA did not, we focused on further characterizing and assessing vaccination via solely MA-BCN formulations.

MA persists while Ag85B does not after vaccination with Ag85B-MA-BCN

Given the lack of knowledge regarding BCN and lipid antigen kinetics, it would be of interest to test whether delivered antigen could persist in the mouse weeks post-vaccination (Fig 3A). We switched from intranasal vaccination to intratracheal (IT) vaccination, as it allowed for the reliable delivery of a larger volume of the vaccine. We found that a significant proportion of the adoptively transferred DN1 T cells could proliferate and become activated in the LNs, lung, and spleen of MA-BCN-vaccinated mice 6 weeks post vaccination (Fig 3B-D). This activation and proliferation were MA-specific, as no activation and proliferation of Ag85B-specific p25 Tg (p25) T cells were observed[34].

Vaccination with MA-BCN leads to antigen persistence 6 weeks post-vaccination.

hCD1Tg mice were IT vaccinated with MA-BCN or BCN. 6 weeks later, CellTrace-labeled p25 and DN1 T cells were adoptively transferred, and T cell activation was measured after 1 week. (A) Experimental design. (B) Representative FACS plots of p25 and DN1 T cells in the LN. Percentage of proliferating (C) and CD44-expressing (D) p25 and DN1 T cells in the LN, lung, and spleen. N = 3 or 6 per condition. Outliers were identified through Grubbs method with alpha=0.05 with one sample removed. Data represented as mean ± SEM. Statistical analysis: 2-way ANOVA. *p < 0.05, **p < 0.01, ***p < 0.001, ****p < 0.0001.

Since BCN allow for the co-loading of lipid and protein antigens, we assessed the ability of dual loaded Ag85B-MA-BCN to activate antigen-specific T cells as well as lead to antigen persistence. We co-loaded Ag85B and MA in BCN and found Ag85B-MA-BCN to have a diameter of 380 nm and PDI of 0.223, comparable to MA-BCN (Fig 4A). Encapsulation efficiency was also relatively high at 70%.Vaccination of hCD1Tg mice with Ag85B-MA-BCN activated and induced proliferation of both p25 and DN1 T cells in the LN, lung, and spleen (Fig 4B-D) at one-week post-vaccination. In addition, DN1 T cells were still able to be activated 6 weeks post-vaccination in the context of an Ag85B-MA-BCN vaccination, while p25 T cells did not show increased activation or proliferation compared to blank BCN control at this time point (Fig 4E-H). Therefore, MA, but not Ag85B, appears to persist in the context of an Ag85B-MA-BCN vaccination.

Ag85B and MA dual loaded BCNs activate antigen-specific T cells without Ag85B persistence.

(A) DLS analysis of blank BCN and Ag85B-MA-BCN. (B-D) hCD1Tg mice were IT vaccinated with Ag85B-MA-BCN or BCN and CellTrace-labeled p25 and DN1 T cells were adoptively transferred the next day. After 1 week, T cell activation and proliferation was measured. (B) Representative FACS plots of p25 and DN1 T cells in the LN. Percentage of proliferating (C) and CD44-expressing (D) p25 and DN1 T cells in the LN, lung, and spleen (N = 4 or 5). (E-G) hCD1Tg mice were IT vaccinated with blank BCN or Ag85B-MA-BCN. 6 weeks later, CellTrace-labeled p25 and DN1 T cells were adoptively transferred, and T cell activation was measured after 1 week. (E) Representative FACS plots of p25 and DN1 T cells in the LN. (F) Percentage of proliferating p25 and DN1 T cells. (G) Percentage of CD44-expressing p25 and DN1 T cells in the LN and lung. (N = 4). Data represented as mean ± SEM. Statistical analysis: 2-way ANOVA. ns = not significant, *p<0.05, **p<0.01, ***p<0.001, ****p< 0.0001.

MA persistence was dependent on encapsulation but not route of delivery or delivery vector

To determine whether BCN encapsulation contributed to MA persistence, we designed an experiment that would allow direct comparison between MA-BCN and free MA. To account for their differential efficiency in stimulating DN1 T cells, we pulsed bone marrow-derived dendritic cells (BMDCs) from hCD1Tg mice with MA or MA-BCN at 10 μg/mL and 5 μg/mL MA concentration, respectively (Fig 5A). hCD1Tg mice were immunized with MA or MA-BCN-pulsed BMDCs at either 1 week or 6 weeks before the adoptive transfer of DN1 T cells, allowing comparison of short-term vs. long-term vaccination conditions. We found that while there were no significant differences in DN1 T cell activation and proliferation between mice vaccinated with free MA and MA-BCN pulsed BMDCs at 1-week post-vaccination, after 6 weeks, DN1 T cell response could only be detected in the MA-BCN vaccination condition (Fig 5B, C). Thus, BCN encapsulation contributed to the persistence of MA. We then tested whether the BCN structure could contribute to antigen persistence by comparing PEG-b-PPS BCNs to PEG-b-PPS MCs using a similar experimental set up (Suppl Fig 3). We found that MA-BCN and MA-MC overall had similar ability to activate DN1 T cells in both short-term and long-term vaccination time points (Suppl Fig 3B, C). Thus, the structure of the encapsulating NP did not play a significant role in antigen persistence of MA.

Route of vaccination or vector type do not affect antigen persistence.

hCD1Tg mice were IT vaccinated with MA or MA-BCN pulsed BMDCs at 6 weeks or 1 week prior to the adoptive transfer of CellTrace-labeled DN1 T cells. T cell activation and proliferation was measured 1 week after adoptive transfer. (A) Experimental design. (B, C) Percentage of proliferating (B) and CD44-expressing (C) DN1 T cells in the LN, lung, and spleen of vaccinated mice. N = 4 per condition. N = 4 per condition. hCD1Tg mice were vaccinated SC (subcutaneously) or IV (intravenously) with blank BCN, MA-BCN, or attenuated Mtb strain and 6 weeks later CellTrace-labeled DN1 T cells were adoptively transferred. (D) Representative FACS plots of DN1 T cells in the LN of mice vaccinated with indicated conditions. (E) Percentage of proliferating DN1 T cells and (F) Percentage of CD44-expressing DN1 T cells in the LN and spleen of mice vaccinated with Blank-BCN (SC), MA-BCN (IV and SC), and attenuated Mtb. N = 3-5 per condition. Data represented as mean ± SEM. Statistical analysis: 2-way ANOVA. ns = not significant, *p<0.05, **p<0.01, ***p<0.001, ****p<0.0001.

To investigate if the IT route of vaccination played a role in the development of antigen persistence, we tested intravenous (IV) and subcutaneous (SC) vaccination routes with MA-BCN. DN1 T cell activation and proliferation were detected at 6-weeks post-vaccination through both routes of vaccination (Fig 5D-F), suggesting that the antigen persistence is not unique to IT vaccination. To assess whether Mtb lipid antigen persistence could also be observed in mice vaccinated with an attenuated strain of Mtb, we vaccinated hCD1Tg mice SC with Mtb mc26206 strain and tested for the persistence of both Ag85B and MA after 6 weeks. The Mtb mc26206 strain displays less virulence in mouse studies than BCG but still confers long-term protection against virulent Mtb challenge equivalent to BCG[35]. We confirmed that no bacterial burden remained at the 6 weeks time point (Suppl Fig 4A). Similar to MA-BCN-vaccinated mice, DN1 T cells were activated and proliferated in the draining lymph nodes and spleen of mice vaccinated with the attenuated Mtb strain (Fig 5D-F). In line with our findings with BCN, no proliferation was seen in p25 T cells (Suppl Fig 4B-D). These results suggest that incorporation of MA in a nanocarrier vaccine allows for the mimicry of the lipid antigen persistence induced by vaccination with an attenuated Mtb vaccine.

MA persists within alveolar macrophages in the lung

To identify the location of antigen persistence, we determined the biodistribution of BCN after IT vaccination using BCN loaded with the hydrophobic dye DiD (DiD-BCN). Since DiD is a lipophilic dye which, like MA, stably loads into the BCN bilayers, we could track the localization of BCN without having to perform additional chemical modifications. We found that IT-administered DiD-BCN were located almost exclusively within the lung at all time points tested (4 h, 24 h, 48 h, 6 days, and 6 weeks) (Fig 6A). We were able to detect fluorescence through both in vivo imaging system (IVIS) and flow cytometry at 6-weeks post-vaccination that was significantly above unvaccinated control (Fig 6A, B), despite DiD quickly losing fluorescence upon mixture into aqueous environments [36], suggesting that some DiD-BCN structure remained intact at this time point. Flow cytometric analysis (Suppl Fig 5) revealed that DiD BCN was primarily found within alveolar macrophages (AMs) at both early and late time points with neutrophils and DCs showing some fluorescence at early time points (4 and 24 h) (Fig 6C).

Persistent MA remains in part due to encapsulation inside alveolar macrophages and activates T cells through DCs.

B6 or hCD1Tg mice were IT vaccinated with BCN loaded with a hydrophobic fluorescent dye (DiD) at 6 weeks, 6 days, 48 hours, 24 hours, or 4 hours prior to the experiment. LN, lung, and spleen were analyzed using (A) In Vivo Imaging System (IVIS). Single cell suspension was then obtained and presence of DiD BCN was quantified in (B) total lung, and (C) CD45− cells, neutrophils (Ly6G+), alveolar macrophages (CD11c+SiglecF+), DCs (CD11c+), monocytes (CD11b+CD11c−), B cells (CD19+), T cells (CD3+), NK cells (NK1.1+), eosinophils (CD11c−SiglecF+) by flow cytometry. Gating strategy exemplified in Suppl Fig 4. hCD1Tg mice were IT vaccinated with either BCN or MA-BCN. After 6 weeks, alveolar macrophages were enriched from lungs using anti-SiglecF. Enriched or flow through cells were co-cultured with DN1 T cells in the presence or absence of hCD1Tg-expressing BMDCs for 48 hours, and DN1 T cells activation was measured. (D) Experimental design. (E) Percentage of CD25-expressing DN1 T cells. Data represented as mean ± SEM. Statistical analysis: 2-way ANOVA. ns = not significant, *p<0.05, **p<0.01, ***p<0.001.

We next tested whether MA likewise persisted primarily within AMs. 6-weeks after vaccination, we enriched for AMs using a column-based magnetic cell isolation system containing SiglecF antibody, which is highly expressed on AMs and a small population of eosinophils in the lung (Fig 6D). We co-cultured enriched AMs or the flow through with or without hCD1Tg-expressing BMDCs and DN1 T cells (Suppl Fig 6). While both enriched AM and flow through fractions from MA-BCN vaccinated mice co-culture with BMDCs could activate DN1 T cells (Fig 6E), co-culture with enriched AMs led to significantly higher DN1 T cell activation, suggesting AMs are the primary location of MA persistence. The ability of flow through to likewise lead to DN1 T cell activation suggests MA may also persist in other cell types or could be attributed to residual AMs in flow through. Furthermore, AMs could not themselves present the MA to DN1 T cells, noted by the lack of T cell activation in the absence of BMDCs, suggesting that MA may be transferred from AMs to DCs for presentation. In fact, CD1b is not expressed on AMs [19, 37] and therefore the presence of CD1b-expressing DCs may be necessary both in vitro and in vivo for T cell activation to occur. These data indicate that MA-BCN mainly persists within AMs in the lung after IT vaccination and CD1b-expressing DCs is required for antigen presentation and activation of MA-specific T cells.

Vaccination leads to DN1 T cell differentiating into T follicular helper-like T cells

To study the memory phenotype of DN1 T cells after MA-BCN vaccination, we constructed a mixed DN1 bone marrow (BM) chimera mouse model (DN1-hCD1Tg), since adoptively transferred DN1 T cells in unvaccinated mice cannot survive long term (Fig 7A). After vaccination, the percent of CD44+ DN1 T cells in the blood increased quickly and remained high until the last time point of 40 days (Fig 7B). Various classical memory T cell subsets have been characterized 1 month after initial antigen-encounter, we therefore determined the DN1 T cell phenotype 6-weeks post vaccination. We found that the most prevalent memory population within DN1 T cells were CD44+CD62L+, markers used to define central memory T cells, particularly in the LN (Fig 7C and Supp Fig 7A). To characterize the memory DN1 T cells, we performed RNA-seq analysis on sorted CD44+CD62L+ (memory) and CD44−CD62L+ (naïve) DN1 T cells from LNs of MA-BCN vaccinated DN1-hCD1Tg BM chimeric mice. We found that memory and naïve DN1 T cells clustered separately after principal component analysis (PCA), despite these samples coming from the same animals (Suppl Fig 7B). A total of 995 differentially expressed genes (DEGs) were identified of which 542 upregulated and 453 downregulated in the memory subset (Suppl Fig 7C). Next, we determined which T cell population memory DN1 T cells most resembled. Toward this end, we obtained data from ImmGen database (including terminally differentiated effector (Te), memory precursor (Tmp), central memory (Tcm), effector memory (Tem), regulatory (Treg))42, and two additional publications (follicular helper (Tfh)43, exhausted (Texh)44) and compared the respective DEG lists. Using PCA, we found that memory DN1 T cells clustered most closely to Tfh, Treg, and Texh (Fig 7D) and when this subset is isolated, most closely to Tfh cells (Fig 7E).

MA-BCN vaccination leads to DN1 T cell differentiation into T follicular helper-like T cells.

Mixed bone marrow chimeric mice were created by adoptive transfer of DN1 bone marrow and congenic CD45.1 hCD1Tg bone marrow to irradiated hCD1Tg mice. After 5 weeks, mice were vaccinated with either MA-BCN or PBS, and DN1 surface marker expression was monitored in the blood and in organs at 6 wks. (A) Experimental design. (B) Percent CD44+ DN1 T cells in the blood at various time points post vaccination. (C) Percent CD44+CD62L+ DN1 T cells in indicated organs at 6 weeks post vaccination. Memory (CD44+CD62L+) and naïve (CD44−CD62L+) DN1 T cells were sorted from MA-BCN vaccinated hCD1Tg-DN1 BM chimeras at 6 weeks post vaccination and subjected to RNAseq analysis. N=3 per condition (D, E) PCA of log fold change of gene expression for each T cell subset was performed relative to internal naïve T cell control. (F) Relative expression of key TFH cell and DEGs in memory subset. (G) Gene enrichment analysis of upregulated DEGs was performed using Metascape. (H) Representative FACS plots of DN1 T cells in the LN of mice. Percentage of (I) CXCR5+PD1+ or (J) KI67+ DN1 T cells DN1 T cells in LN, lung, and spleen. N = 3-5 per condition. Data represented as mean ± SEM. Statistical analysis: 2-way ANOVA. *p<0.05, **p<0.01, ****p < 0.0001.

Within the DEGs, we noted upregulation of key Tfh cell transcription factors BCL6 and TCF1 (TCF7), chemokines CXCR5 and CXCR4 and surface receptor ICOS, important for Tfh cell migration into the germinal center, and adaptor protein SAP (SH2D1A), important for T cell-dependent B cell immunity (Fig 7F)45. Inhibitory receptors PD-1 (PDCD1), TIGIT, CTLA4, and transcription factor EOMES, which are known to be expressed on Tfh cells were also upregulated. In addition, we observed downregulation of transcription factor KLF2, which can inhibit Tfh cell differentiation, and CCR7 and IL-7R, receptors known to have decreased expression in TFH cells. Consistent with our transcriptome analysis, we found increased percentage of CXCR5+PD-1+ DN1 T cells in MA-BCN vaccinated mice 6 weeks post-immunization (Fig 7G, H). Both our transcriptomic and experimental data highlighted that memory, but not naïve DN1 T cells, were proliferating, given the enrichment for the mitotic cell cycle process (Fig 7I) and increased KI67+ percentage within this population (Fig 7J). MA-BCN vaccination thus led to the differentiation of DN1 T cells into a proliferating Tfh-like T cell population.

Discussion

Research on the development of subunit vaccines containing lipid antigens is limited. This is in part due to a lack of research into appropriate lipid vaccine formulations and their associated properties. Some early evidence suggests that lipid vaccines could contribute to protection in Mtb challenge in guinea pig model [38, 39], which naturally express group 1 CD1 molecules [40]. Vaccination with a formulation composed of total Mtb lipid encapsulated within liposomes with either dimethyl-dioctadecyl-ammonium (DDA) and/or QS-21 adjuvants showed lower bacterial burden and pathology in lung and spleen compared to unvaccinated controls [38]. In another study, two Mtb lipids (diacylated sulfoglycolipids and the phosphatidylinositol dimannosides) encapsulated into liposomes formulated with DDA and trehalose dibehenate led to a decreased bacterial burden in the spleen and overall improved pathology in lung and spleen [39]. This work motivated our current study, where we used hCD1Tg mice that express human group 1 CD1 molecules, including CD1b and MA-specific (DN1), as well as p25-specific T cells to investigate the use of nanocarriers coloaded with both Mtb lipid and protein antigens during vaccination.

We found that PEG-b-PPS BCN could effectively encapsulate MA as well as the protein antigen Ag85B to elicit activation of antigen-specific T cells in vaccinated mice (Fig 4). To our knowledge, this is the first time a dual protein and lipid loaded nanocarrier has been shown to effectively stimulate both lipid and peptide antigen-specific T cells simultaneously in vivo. PEG-b-PPS BCN can encapsulate both hydrophobic and hydrophilic molecules, making it an excellent platform for subunit vaccine delivery as both antigens and adjuvants can be co-loaded and delivered to the same cellular targets. Prior work on TB vaccines has shown that the inclusion of several protein antigens spanning the Mtb life cycle along with adjuvants with multiple boosts is paramount in the development of a vaccine that can provide effective protection against Mtb [30, 41]. As we have established that PEG-b-PPS BCN can easily incorporate a variety of payloads, future work will involve determining the optimal antigen and adjuvant combination for providing protection in a Mtb challenge experiment.

As several formulations of PLGA-NP have been approved for clinical uses [42], we compared MA encapsulated by either BCN or PLGA-NP and found that while PLGA appeared to offer greater, yet less controllable stimulation in vitro (Fig 2), BCN exhibited significantly greater efficacy in vivo (Fig 2). The result is not completely unexpected as a more fragile PLGA-NP could allow for fast intracellular release and antigen presentation in vitro while also leading to significant antigen loss prior to APC interaction in vivo [43]. Previous studies have documented PLGA’s innate pro-inflammatory effects, while other studies have highlighted the absence and suppression of inflammatory response [44–46]. In contrast and as previously observed for PEG-b-PPS nanocarriers [23], we showed that unloaded BCN have little inflammatory effect (Fig 2), allowing their immunomodulation to be dictated by the properties of the loaded antigen and adjuvant. Thus, PEG-b-PPS is a promising polymer for the formulation of subunit vaccines.

We proceeded with a pulmonary (IN or IT) vaccination approach in this study based on emerging evidence that mucosal vaccination can induce a stronger immune response in the lung against Mtb in humans [47] and improve protection following vaccination with nanocarriers in mice [48]. We found that in the pulmonary route of vaccination, BCN primarily were taken up and remained within alveolar macrophages in the lung (Fig 6). In fact, the ability of nanocarriers to target alveolar macrophages or other APCs may contribute to their distinct capacity for inducing antigen persistence, which was not seen with cellular carriers such as MA-pulsed BMDCs (Fig 5). Alveolar macrophages are the first line of defense of the lungs, responsible for keeping alveoli sterile by taking up inhaled particles and pathogens through phagocytosis [49–51]. Targeting these cells may be particularly beneficial in a TB vaccine design since they may also be the main site of Mtb latency [52]. While MA remained within alveolar macrophages in the lung, T cell activation appeared to occur prominently within the draining lymph nodes (Fig 2), suggesting that either MA was transported from the lungs to lymph nodes where the presentation took place or T cells activated in the lung migrated to the lymph nodes. The first option seems most likely, as CD1b-expressing DCs are more prevalent in the lymph nodes [53]. Furthermore, in Mtb infection models, it has been previously shown that apoptotic macrophage and vesicles could deliver mycobacterial antigens to DCs that then activate cognate T cells [54, 55]. Similarly, we showed that alveolar macrophages from MA-BCN vaccinated mice could only lead to T cell activation in the presence of BMDCs from hCD1Tg mice (Fig 6).

We demonstrated MA persistence in the lung 6 weeks post-vaccination using an in vivo antigen presentation assay with adoptive transfer of naïve MA-specific T cells. Sustained release of antigen through vaccination is thought to be key in the development of a robust adaptive immune response [56, 57]. Recently, spike protein was found within the draining lymph nodes of vaccinated human subjects 60 days after COVID-19 vaccination with mRNA-1273 or BNT162b2 [58]. In mice, pulmonary vaccination with adenovirus expressing influenza nucleoprotein supported long term maintenance of CD8 TRM (resident memory) cells [59]. Unlike in this work, antigen persistence did not occur after subcutaneous vaccination, suggesting a distinct mechanism of antigen archiving. In a different vaccine model, subcutaneous administration of ovalbumin paired with adjuvants polyl:C and anti-CD40 led to antigen archiving by LECs [11, 12]. Similar to our findings, antigen-specific T cell activation required cross-presentation by DCs [11, 60]. Both LECs and FDCs retain antigens within a non-degradative compartment [11, 61], suggesting that residual antigens in our model may similarly remain in the endosome within undigested BCN. We have previously demonstrated the enhanced stability of BCN within cell endosomes in vitro for extensive periods of time [22]. This possibility is corroborated by the finding that DiD, which is easily quenched in aqueous solution [36], remained fluorescent 6 weeks post-vaccination (Fig 6). The lack of persistence for Ag85B could be due to its leakage out from the aqueous pores of BCN, which we have observed for other proteins [29]. Peptide antigens may be modified to include a hydrophobic tail for stable retention within BCN and therefore allow for peptide antigen persistence as well. MA’s extreme hydrophobic surface, which correlates with its low permeability as a part of Mtb cell wall, may also contribute to its prolonged persistence in vivo [62].

After infection and vaccination, peptide antigen persistence has been shown to occur through archiving by binding to CD21 receptors on follicular dendritic cells (FDCs) in lymph nodes [3, 4]. MA antigen persistence may likewise be a natural phenomenon that can occur after infections, although its detection may be difficult given the lack of available assays to detect very small amounts of lipid antigen. T cells may, in fact, be more sensitive than any commercially available tool. While we determined that MA persists within alveolar macrophages after pulmonary vaccination, the exact location of persistence after SC, IV, or attenuated Mtb vaccinations remains to be tested. Overall, MA antigen persistence is a robust phenomenon which occurs independently of the vector and vaccine route of administration, and thus may play a role in the efficacy of a variety of vaccine strategies. Additional work will focus on whether lipid antigen persistence is found across the lipid family or specific to lipids with certain properties.

The use of memory T cells in the setting of antigen persistence may be debatable, given that memory response has historically referred to the adaptive immune response in the setting of antigen clearance [63]. However, given that the persistence of MA occurs even in the setting of attenuated Mtb vaccination, Mtb lipid-specific long-term adaptive immunity will likely naturally exist in an antigen persistent environment. We found that DN1 T cells in this setting expressed key markers of Tfh cells such as CXCR5, PD-1, BCL6, and TCF1 (Fig 7). T cells differentiate into Tfh cells through the expression of master transcription factor regulator BCL6, which directs the upregulation of chemokine receptors CXCR5 and CXCR4 and thus migration into the germinal center[64, 65]. In the germinal center, Tfh cells provide help to B cells for affinity maturation and class switching. Other unconventional T cells are known to behave similarly such as CD1d-restricted NKT cells, which can acquire a Tfh phenotype and provide B cell help through both cognate and non-cognate interactions[66]. Additional work will look at whether DN1 or other group 1 CD1-restricted T cells can functionally act as Tfh cells to provide B cell help, as may be suggested by the existence of IgG antibodies against MA and glycolipid lipoarabinomannan in TB patients [67–69].

While persistence of protein antigen has been shown to improve the quality of protective immunity [7, 11, 70, 71], it is not yet known how lipid antigen persistence may affect the protective efficacy of lipid-specific T cells. At the very least, antigen persistence increases the probability of T cell/DC encounters by increasing the time that the antigen is available. Additional work will need to address the function of lipid-specific T cells differentiated in antigen persistent environments. As the ultimate goal of this work is to improve the efficacy of existing subunit vaccines, future work will assess the protective capacity of lipid antigen vaccines paired with additional protein antigens and adjuvants.

Materials and Methods

Ethics statement

This study was carried out in accordance with the recommendations in the Guide for the Care and Use of Laboratory Animals of the National Institutes of Health. The protocol was approved by the Institutional Animal Care and Use Committee of Northwestern University (Protocol number: IS000011717).

Poly (ethylene glycol)-block-poly (propylene sulfide) polymer synthesis

Poly (ethylene glycol)17-block-poly-(propylene sulfide)75 (PEG17-b-PPS75) copolymer was synthesized as previously reported [72, 73]. Briefly, monomethoxy PEG (molecular weight=750 Da) was functionalized with mesyl chloride and substituted with thioacetate to produce a PEG acetate initiator for the polymerization for PPS.

Bicontinuous nanosphere fabrication via flash nano precipitation

BCN were produced using FNP technique as previously described[72]. In brief, PEG17-b-PPS75 BCN polymer and any hydrophobic cargo was dissolved in THF and loaded onto a syringe. The hydrophilic cargo was loaded onto a separate syringe, and both impinged against one another through the confined impingement mixer into scintillation vial. The vial was placed in a vacuum desiccator overnight to remove the organic solvent. NPs were filtered through LH-20 column with PBS as an eluent and suspended in PBS for later use. MA encapsulation efficiency was analyzed using coumarin derivatization as previously described [21]. MA was loaded at a concentration of 100 µg/mL of total BCN formulation and DiD was loaded at 5 µg/mL total BCN formulation. MA loading capacity in BCNs was 2%. To measure the encapsulation efficiency of Ag85B, Ag85B-MA-BCN were centrifuged at 10,000 g for 10 minutes. The supernatant was collected and concentrated using Amicon® Ultra-4 10K centrifugal filter devices. To determine the unencapsulated protein concentration, the supernatant was incubated with Pierce™ BCA Protein assay reagent following the manufacturer protocol and the absorbance was measured at 660 nm using a SpectraMax M3 multi-mode microplate reader (Molecular Devices, LLC). The percentage encapsulation efficiency was calculated as a percent of encapsulated protein to the total amount of protein added. Ag85B was encapsulated at a concentration of 140 µg/mL and loading capacity was found to be 2.88%.

Poly (lactic-co-glycolic acid nanoparticle (NP) fabrication

PLGA-NP were prepared by the Oil-in-Water single emulsion method. Briefly, PLGA (20 mg in 800 µL dichloromethane) organic phase was emulsified using an ultrasonic processor with aqueous phase containing 2 mL of polyvinyl alcohol (PVA) solution (2.5% w/v) to form an emulsion. The emulsion was then added dropwise into 4 mL of stirring 0.5% w/v PVA solution at room temperature to evaporate the organic solvent. NPs were collected after 4 hours of stirring followed by centrifugation at 10,000 x g for 10 minutes. After centrifugation, NPs were washed twice with cold water to remove residual PVA and redispersed in PBS. Hydrophilic (Texas red-Dextran) or hydrophobic (MA) loading was performed by adding the payloads to the aqueous and organic phases, respectively.

Nanocarrier characterization

The nanocarrier size (z-average diameter) and polydispersity was measured using dynamic light scattering (DLS) using a Nano 300 ZS Zetasizer (Malvern Panalytical, Malvern, U.K.). BCN and PLGA morphology was visualized with cryogenic transmission electron microscopy (cryoTEM) as previously detailed [21]. Transmission electron microscopy (TEM) was performed to image BCN using uranyl acetate as a negative stain as previously reported [22]. BCN aggregate structure and internal morphology was characterized with small angle X-ray scattering (SAXS). These studies were performed at the DuPont-Northwestern-Dow Collaborative Access Team beamline at Argonne National Laboratory’s Advanced Photon Source with 10 keV (wavelength λ = 1.24 Å) collimated X-rays, as described previously[72].

Confocal imaging

Confocal images of RAW 264.7 macrophages stained with lysosomal dye Lysotracker™ (green) and NucBlue™ stain (blue) following incubation with Texas Red-labeled PLGA-NP and BCN (red) for 8 hr. The cells were then imaged within a humidified chamber using a 63X oil-immersion objective on a SP5 Leica Confocal Microscope using HyD detectors and lasers. Data analysis was performed using ImageJ software.

Mouse strains

Human CD1 transgenic mice (hCD1Tg) in C57BL/6 background or MHC II-deficient background [19] and CD1b-restricted MA-specific TCR transgenic mice on Rag−/− background (DN1Tg/hCD1Tg/Rag−/−) [20] were generated in our lab as previously described. P25-specific TCR transgenic mice (Jackson lab) were crossed onto Rag−/− background. Both males and females were used.

Antibodies and flow cytometry

For cell surface staining, cells were pre-incubated with 2.4G2 Fcγ RII/RIII blocking mAb for 15 minutes and then stained with the appropriate combinations of mAbs diluted in HBSS + 2% FBS for 30 minutes at 4°C. Cells were analyzed on a BD FACS CantoII, or Cytek Aurora Spectral Cytometer and data were processed with FlowJo software (TreeStar). The complete list of antibodies use is as follows: anti-human Vβ5.1 (LC4, Invitrogen), anti-TCRβ (H57-597, BioLegend), anti-CD44 (IM7, BioLegend), anti-Vβ11 (KT11, BioLegend), anti-CD45.2 (104, BioLegend), anti-CD45.1 (A20, BioLegend), anti-SiglecF (E50-2440, BD Bioscience), anti-CD45 (30-F11, BioLegend), anti-CD11c (N418, BioLegend), anti-CD11b (M1/70, BioLegend), anti-Ly6G (1A8, BioLegend), anti-CD19 (6D5, BioLegend), anti-NK1.1 (PK136, BioLegend), anti-CD3 (17A2, BioLegend), Zombie Red (BioLegend), anti-mouse CD185/CXCR5 (L138D7, Biolegend), anti-mouse CD279/PD-1 (29F.1A12, Biolegend), and anti-Ki67 (SolA15, Invitrogen).

In vitro nanoparticle and lipid titration with MA-specific DN1 T cells

Nanoparticle and lipid titrations were performed as previously described [21]. In brief, hCD1Tg MHC II−/− bone marrow-derived dendritic cells (BMDCs) were pulsed with varying quantities of nanoparticle of sonicated MA (Sigma-Aldrich, St. Louis, MO) and then co-cultured for 48 hours with DN1 T cells isolated from lymph nodes (axial, brachial, submandibular, inguinal) of DN1Tg/hCD1Tg/Rag−/− mice. The supernatant was used for IFN-γ enzyme-linked immunosorbent assay (ELISA) performed as described previously [21] and cells were stained for activation.

In vitro nanoparticle and lipid titration with human cells via ELISPOT assay

Hydrophobic polyvinylidene fluoride 96-well plates (Millipore, Bedford, MA) were coated with anti-human IFN-γ mAb 1-D1K (Mabtech, Cincinnati, OH) diluted 1:400 in PBS (Gibco, Waltham, MA) and incubated at 4°C overnight. The following day, mAb was removed and each plate was washed and incubated at room temperature for 2 hours in the presence of media. Human monocyte-derived dendritic cells and MA-specific M11 T cells were seeded 50,000 cells and 2,000 cells per well, respectively, together with serially diluted MA and NPs. After 16-hour incubation, plates were washed and incubated with biotinylated anti-human IFN-g (7-B6-1), diluted 1:3000 in PBS + 0.5% FBS, for 4 hours. Plates were then washed with PBS 5 times and incubated for 2 hours with ExtrAvidin-Alkaline phosphatase diluted 1:1000 in PBS + 0.5% FBS. Spots were visualized after incubation with the BCIP/NBT substrate for up to 20 minutes and then imaged with ImmunoSpot reader v.2.0.

Mouse vaccination

For both intranasal (IN) and intratracheal (IT) vaccination, mice were anesthetized using inhaled isoflurane. For IN vaccination, 25 µL of liquid was administered at the nostrils of the mice for a total of 1 μg of MA per mouse. For IT vaccination, mice were administered 25 or 50 μL of liquid for a total of 2.5 μg of MA with or without 2.0 μg of Ag85B. For BMDC vaccinations, hCD1Tg BMDCs were cultured in GM-CSF and IL-4 for 5 days[74] and pulsed with MA-BCN (5 μg/mL of MA) or free MA (10 μg/mL of MA) for 24 hours. Cells were washed and resuspended in PBS and IT vaccinated at a dose of 1×106 BMDCs/mouse. Intravenous vaccination was administered by tail vein injection. Subcutaneous vaccination was administered between the shoulders over the neck portion of the mouse.

Vaccination with attenuated Mtb

Attenuated Mtb strain (H37Rv ΔpanCD ΔleuCD referred to as mc2 6206) was kindly provided by Jacobs’ Laboratory at Albert Einstein College of Medicine. Bacteria were grown as previously described in supplemented Middlebrook 7H9 media [35, 75]. Mice were vaccinated subcutaneously with 1×106 CFU in 100 μL of PBS. Mycobacterial burden quantification was performed by plating serial dilutions of lung or spleen homogenate on Middlebrook 7H11 agar plates [35, 75].

T cell Adoptive transfer and cell preparation

DN1 or p25 T cells were isolated from lymph nodes of DN1Tg/hCD1Tg/Rag−/− or P25Tg/Rag−/− mice, respectively. Cells were labeled with CellTrace™ Violet reagent (Invitrogen™) as per manufacturer’s instructions and 3-5×106 cells were injected into hCD1Tg mice either 1 day before vaccination (short term) or 6 weeks post-vaccination (long term). On day 7 post-adoptive transfer, draining lymph nodes (axial, brachial, mediastinal), lung, and spleen were obtained. Single-cell suspensions were prepared by mechanical disruption in HBSS/2% FBS. The lung was digested with collagenase IV (1mg/ml; Sigma-Aldrich) and DNase I (30µg/ml; Sigma-Aldrich) for 30 minutes at 37°C before the disruption.

Biodistribution imaging

BCN were loaded with DiD dye (Thermo Fisher) and administered to mice IT at 4 hours, 24 hours, 48 hours, 6 days, and 6 weeks prior to the final time point, at which all mice were sacrificed and analyzed together. Lung, spleen, and draining lymph nodes were harvested and imaged using a near-IR in vivo Imaging System (IVIS; Center for Advanced Molecular Imaging, Northwestern University). The single-cell suspension was then prepared from indicated organs and cells were stained for flow cytometric analysis.

Alveolar macrophage enrichment and T cell co-culture

hCD1Tg mice were immunized IT with either BCN or MA-BCN (2.5 μg of MA/mouse). 6 weeks later, lungs were isolated and single-cell suspension was stained with anti-SiglecF-PE (E50-2440, BD Bioscience). SiglecF-positive cells were enriched using anti-PE MultiSort Kit (Miltenyi Biotec). 1×105 flow through and AM enriched fractions were co-cultured with 1×105 hCD1Tg BMDCs and 3×105 DN1 T cells for 48 hours and cells were stained for the expression of activation markers.

Generation and vaccination of BM chimeras

Bone marrow cells from DN1Tg/hCD1Tg/Rag−/− (CD45.2) and hCD1Tg mice (CD45.1) were depleted of mature T cells using anti-Thy-1.2 (AT83.A-6) plus rabbit complement (Cedarlane Laboratories). Equal numbers of cells (5×106) were adoptively transferred into hCD1Tg (CD45.1) mice irradiated with 900 rads. After 6 weeks, BM chimeric mice were intratracheally vaccinated with either 2.5 μg of MA in MA-BCN or equivalent volume of PBS and the activation status of DN1 T cells were monitored in the blood 1-2 times per week until experiment completion.

RNA-Seq analysis

CD44+CD62L+ and CD44−CD62L+ DN1 T cells were sorted from LNs of DN1-hCD1Tg BM chimera mice vaccinated with MA-BCN at 6 weeks post vaccination by BD FACS Ariax with purity >98%. RNA was extracted with RNAeasy® mini kit (Qiagen). Libraries were generated by using the Illumina Truseq preparation kit and sequenced on HiSeq4000. Reads were analyzed with Ceto pipeline (https://github.com/ebartom/NGSbartom) using STAR and HTseq for alignment on mm10 mouse genome and reading counting, as described previously [76]. Paired differential expression analysis was performed using DESeq2 to account for original mouse sourcing[77]. All downstream analysis was performed in R and figured generated using ggplot. Gene enrichment analysis was performed using Metascape [78]. Comparisons to other T cell profiles were performed using raw transcriptome data from ImmGen database[79], and two additional publications for follicular helper (TFH)[80], exhausted (Texh)[81]. Analysis was performed as described above to obtain gene expression fold change values relative to internal naïve condition.

Data accessibility

The data discussed in this publication have been deposited in NCBI’s Gene Expression Omnibus and are accessible through GEO Series accession number GSE226075 (https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226075).

Statistical analysis

Statistical analyses were performed using Prism software 9 (GraphPad software, Inc.). Multi-group comparisons were done using two-way analysis of variance (ANOVA) with significance determined by Tukey’s multiple comparison. Where appropriate, outliers were identified through Grubbs method with alpha=0.05. Statistically significance denoted as ns = not significant, *p<0.05, **p<0.01, ***p<0.001, and ****p< 0.0001.

eLife Assessment

In the present study, authors generate a new formulation built upon a previous nanoparticle platform to generate a new system termed bicontinuous nanospheres (BCN), allowing for the dual incorporation of lipid and protein antigens. Authors generate mycolic acid (MA)-loaded BCN perform a series of characterization studies to demonstrate the superior performance of this new formulation relative to the original one in term of antigen persistence, a quality needed to sustain responses after vaccination. This work provides important new insights relevant to the TB vaccine field and it suggest that alternative antigens to proteins could be used in TB vaccine formulations. The data is convincing and will be interesting to individuals working on tuberculosis, vaccines and basic immunology.

Reviewer #1 (Public Review):

It is well established that tuberculosis (TB), which is caused by Mycobacterium tuberculosis (Mtb), is a leading cause of mortality and morbidity worldwide. However, the only vaccine licensed against tuberculosis is Bacille Calmette Guerin (BCG), has been around for nearly a century, and has limited efficacy in adults. Herein, the authors sought to investigate the effectiveness of a nanoparticle-based formulation of a subunit vaccine composed of Mtb lipid and protein antigens. The authors found that they were able to load the lipid, mycolic acid, into their nanoparticles without disrupting the architecture, and that the loaded particles activated T cells both in vitro and in vivo. Moreover, when they vaccinated with particles loaded with both lipid and protein antigens, they found that the lipid antigen persisted, and mycolic acid-specific T cells were able to be activated 6 weeks post-vaccination, in contrast to peptide-specific T cells. The authors investigated further and found that persistence required the nanoparticle encapsulation, rather than free lipid, and that it was independent of route (intratracheal, intravenous, or subcutaneous) of administration. To address the mechanisms underlying antigen persistence, the authors loaded the nanoparticles with a dye and demonstrated that the nanoparticle encapsulated lipid antigen was primarily stored in lung alveolar macrophages and that CD1b+ dendritic cells presented the antigen to mycolic acid specific T cells. Finally, the authors conducted mixed bone marrow chimera studies to examine the phenotype of the mycolic acid specific T cells and found that the memory T cell population phenotypically resembled T follicular helper, regulatory T cells, and exhausted T cells. Interestingly, while a large percentage of these lipid antigen specific T cells in the lymph nodes, lung and spleen were CXCR5+PD1+, the cells were still proliferating (Ki67+). Overall, this is a comprehensive study that has the potential to significantly enhance the field.

Reviewer #2 (Public Review):

The work presented here by Morgun et al is performed in the context of vaccine development, a field especially active in the context of tuberculosis (TB). The generation of a new vaccine either enhancing or replacing the 100-year-old BCG is urgently needed.

Most subunit vaccines integrate protein antigens formulated with adjuvants and there are few examples on the performance of subunit vaccines integrating lipid antigens. Considering the hydrophobic and lipid nature of the mycobacterial cell envelope studies assessing the suitability of mycobacterial lipids in vaccine formulations may contribute to generate new vaccines to tackle the disease.

The mycobacterial lipid antigens under study are mycolic acids (MA), which are located at the cell wall covalently linked to arabinogalactan. These lipids carry extremely long chain fatty acids of up to 60-90 carbons.

The group has previously shown that formulating MA into micellar nanocarriers and vaccinating mice intranasally it could activate CD1-restricted T cells. However, this formulation did not allow for the incorporation of protein antigens.

This work is novel, and it brings new data of high relevance for the TB vaccine field pointing to alternative formulations and antigens and immune mechanisms.

Authors assay different routes of vaccination but the main results are obtained using non-conventional vaccination routes. Although, it maybe out of the scope of the paper, no protection studies are provided.

Several recommendations are given to improve the quality and the readability of the manuscript.

1. Authors elaborate the introduction solely highlighting the relevance of antigen persistence in the context of vaccination. However, it is well known that several mycobacterial antigens (Lipids and proteins) can cause detrimental responses when overexposed to the immune system. In this regard, it would be appropriate to introduce the possibility of the occurrence of exhaustion when prolonged exposure to antigens is happening, which is the main theme of this paper.

2. Authors need to provide more information about the source of MA. It is briefly mentioned in the materials and methods section that it was obtained from Sigma. If that is the case, it would be ideal to show the integrity of the polysaccharide in term of balance and abundance between different MA species.

3. Building up on the previous comment, MA is a complex mixture of polysaccharides including multiple lengths of fatty acids and modifications. Could the authors comments on the potential variability of MA structure and potential impact on immune responses?

4. How do the authors explain the lack of stimulation of cell proliferation induced by MA-PLGA formulation? Does this result contradict previous findings?

5. Fig 3. Authors switch to IT administration simply arguing against the limitation of IN delivery regarding its low volume. However, administration via IN could be done in an iterative manner. According to this change, this reviewer asks whether the performance of MA-PLGA could now be comparable to BCN-MA using IT instead.

6. What would be the reasons of the no role of encapsulating NP in the persistence of MA?

7. Authors need to discuss to what extent the MA location into AM is route dependent.

8. Also, AM are programmed to sustain low immune responses because of their unique location in the lung. In fact, Mtb uses this to replicate while immune response is mounted. In this regard, accumulation of MA into this compartment may not be relevant for the overall immune response. In other words, what would be the contribution of this population to the T cell activation?

9. Could the T cells responses measured be due to the reduced fraction of DC loaded with BCN-MA at initial time points?

Author Response

The following is the authors’ response to the original reviews.

We would like to thank both reviewers and the editor for their time and effort in carefully reviewing and comprehending our manuscript. We are grateful for their thorough assessment, as well as the insightful questions and suggestions they have provided. We have taken into account the questions and comments raised by the reviewers, and we have incorporated the necessary revisions accordingly. In the following pages the reviewers’ comments are italicized. Our replies are in normal script.

In addition to revisions suggested by reviewers we also added a new summary schematic (Fig 8) and minor changes to acknowledgments.

Reviewer 1

This is a very strong study with few concerns. Regarding DN1+ T cell function, the authors assessed IFN-γ and activation markers, but it is unclear if the cells are polyfunctional (produced high levels of other cytokines at 6 weeks) or if there were changes in the humoral response (serum Ab titers or size/ number of germinal centers.)

Thank you for your thorough assessment of our work and your kind comments.

a. We observed a decreased IFN-γ and TNF-α production in antigen experienced DN1 T cells compared to naïve DN1 T cells, which is consistent with findings in Tfh cells.

b. We tested for anti-MA IgM and IgG production but did not observe an increase in these antibodies in the vaccinated setting. It is possible that additional inflammatory stimulation, such as from an adjuvant or infection, may be necessary to trigger sufficient antibody level for detection using ELISA.

c. We did not measure the number or size of germinal centers in this study, but future investigations could explore this aspect.

Reviewer 2

1. Authors elaborate the introduction solely highlighting the relevance of antigen persistence in the context of vaccination. However, it is well known that several mycobacterial antigens (Lipids and proteins) can cause detrimental responses when overexposed to the immune system. In this regard, it would be appropriate to introduce the possibility of the occurrence of exhaustion when prolonged exposure to antigens is happening, which is the main theme of this paper.

Thank you for bringing these points to our attention. We have added a paragraph in the discussion section (page 15-16, line 372-386), addressing the implications of our findings in relation to exhaustion in the context of antigen persistence during chronic viral infections. We have also provided an example involving the lipid trehalose 6,6’-dibehenateled (TDM), a known virulence factor for Mtb, which has been utilized in several subunit vaccines without demonstrating significant toxicity.

1. Authors need to provide more information about the source of MA. It is briefly mentioned in the materials and methods section that it was obtained from Sigma. If that is the case, it would be ideal to show the integrity of the polysaccharide in term of balance and abundance between different MA species.

We obtained M. tuberculosis MA from Sigma, which comprises α-, keto-, and methoxy MA forms with an average combined lipid tail length of 80 carbons. MA-specific T cells preferentially recognize these three forms of MA have been identified in humans. We have provided more detailed information regarding the MA in the Materials and Methods section (page 17, line 429-431).

1. Building up on the previous comment, MA is a complex mixture of polysaccharides including multiple lengths of fatty acids and modifications. Could the authors comments on the potential variability of MA structure and potential impact on immune responses?

The binding capacities of Group 1 CD1-restricted T cells can be influenced by various factors, including specific head groups, lipid tail length, and structure of the lipid tail. Notably, DN1 T cells have been shown to have higher binding affinities towards keto and methoxy MA, while displaying weaker binding to α-MA (Van Rhijn et al., 2017, Eur. J. Immunol. 47:1525). In our study, we successfully utilized a mixture of MA to activate DN1 T cells, indicating that the required subtypes of MA were present in sufficient quantities to elicit this activation. In future investigations focusing on the polyclonal immune response, incorporating a mixture of MA and possibly other Mtb lipid antigens will enable a broader spectrum of T cell activation. This, in turn, is expected to enhance the overall effectiveness and robustness of protection in challenge experiments.

1. How do the authors explain the lack of stimulation of cell proliferation induced by MA-PLGA formulation? Does this result contradict previous findings?

This study represents the first instance of utilizing PLGA as a delivery system for a lipid antigen via a pulmonary vaccine route, despite its previous applications in numerous other vaccine formulations. Therefore, we do not think our findings contradicts any existing research in the field. It is worth noting that the immunogenicity of PLGA can be influenced by the specific polymer chemistry and formulation, which may account for potential variations in the observed effects. We have added additional text to the discussion (page 13, line 310 – 313) to address this point.

1. Fig 3. Authors switch to IT administration simply arguing against the limitation of IN delivery regarding its low volume. However, administration via IN could be done in an iterative manner. According to this change, this reviewer asks whether the performance of MA-PLGA could now be comparable to BCN-MA using IT instead.

PLGA possesses an inherent background adjuvant effect, which may not be ideal for precisely stimulating group 1 CD1-restricted T cells, as a considerable proportion of these T cells exhibit some level of autoreactivity (Li, et al, 2011, Blood 118:3870, De Lalla et al., 2011, Eur. J. Immunol. 41:602; de Jong et al, 2010, Nat. Immunol. 11:1102). Notably, our observations revealed that blank PLGA-NP exerted a significant stimulatory effect on both mouse (DN1) and human (M11) MA-specific T cells (Fig. 2A-D). This underscores the advantage of the BCN system, which lacks detectable adjuvant effects and enables a more controlled, dose-dependent augmentation of T cell responses with increasing concentrations of loaded MA. Therefore, we did not further evaluate the impact of PLGA-MA using the IT route of vaccination.

1. What would be the reasons of the no role of encapsulating NP in the persistence of MA?

In this study, we have provided evidence to support the notion that encapsulation plays a role in antigen persistence, as demonstrated in Fig. 5A-C. Specifically, we directly compared the persistence of MA when delivered encapsulated in BCNs versus without encapsulation in BCNs, using DC pulsing and IT vaccination as the delivery methods. Our results indicate that at 6 weeks post-vaccination, MA encapsulated in BCNs can activate DN1 T cells, while free MA does not. These findings may initially appear to be contradictory to those depicted in Fig. 5D-F, where antigen persistence is observed following vaccination with attenuated Mtb. However, we propose that the attenuated Mtb bacteria may function similarly to nanoparticles by encapsulating and containing MA, thereby facilitating its persistence within the host. We appreciate the opportunity to clarify these points (page 15, line 364-367). Encapsulation within PEG-PPS NP may also contribute to two additional mechanisms. First, we have demonstrated that PEG-PPS NPs target myeloid cell populations (Burke et al., 2022, Nat. Nano. 17:319), such as alveolar macrophages, that can serve as antigen persistence depots as well as present CD1b/MA complexes on their surfaces. NPs allow more efficient delivery to these cells, whereas otherwise the lipid would bind to albumin, HDL, LDL, and other lipid carriers in blood for a broader, non-specific biodistribution, which would include cells less efficient at antigen persistence or presentation. Second, we previously demonstrated that the BCN nanostructure is highly stable within cells, supporting a slow intracellular release (Bobbala et al., 2020, Nanoscale 12:5332). This could assist with a more sustained presentation of lipid antigen by targeted cells in contrast to free form lipid or NPs (like PLGA) that rapidly degrade within cells. Indeed, low levels of fluorescently tagged BCNs were still detectable 6 weeks post-vaccination (Fig. 6B). Our future studies will further investigate this hypothesis.

1. Authors need to discuss to what extent the MA location into AM is route dependent.

The localization of MA within alveolar macrophages (AMs) in the lung is likely specific to intratracheal (IT) vaccination. Therefore, mice vaccinated subcutaneously (SC) or intravenously (IV) may possess distinct antigen persistence depots. We have made modifications to the discussion section to further emphasize this point (page 15, line 359-364).

1. Also, AM are programmed to sustain low immune responses because of their unique location in the lung. In fact, Mtb uses this to replicate while immune response is mounted. In this regard, accumulation of MA into this compartment may not be relevant for the overall immune response. In other words, what would be the contribution of this population to the T cell activation?

It is likely that AMs primarily function as antigen depots and do not directly contribute to the activation of DN1 T cells. This assertion is supported by our findings, as co-culturing AMs with DN1 T cells alone did not result in T cell activation (Fig. 6E). However, we observed that the presence of hCD1Tg-expressing bone marrow-derived dendritic cells was necessary for DN1 T cell activation in vitro, which likely reflects a similar phenomenon occurring in vivo.

1. Could the T cells responses measured be due to the reduced fraction of DC loaded with BCN-MA at initial time points?

Regarding the T cell response observed in Fig. 5A-C, where we used DCs to deliver either free MA or MA-BCN, we took steps to address potential differences in loading capacity between the two at initial time points. Specifically, DCs were pulsed with a concentration of 10 𝜇g/mL for free MA and 5 𝜇g/mL of MA-BCN (the figure legend has been modified to clarify this point, page 37, line 962 - 963). To ensure approximate equivalence in loading, we examined the immune response one week after vaccination and found no statistically significant difference between the two methods.

Version 2

Abstract

Abstract

Mycobacterium tuberculosis (Mtb) infection elicits both protein and lipid antigen-specific T cell responses. However, the incorporation of lipid antigens into subunit vaccine strategies and formulations has been underexplored, and the characteristics of vaccine-induced Mtb lipid-specific memory T cells have remained elusive. Mycolic acid (MA), a major lipid component of the Mtb cell wall, is presented by human CD1b molecules to unconventional T cell subsets. These MA-specific CD1b-restricted T cells have been detected in the blood and disease sites of Mtb-infected individuals, suggesting that MA is a promising lipid antigen for incorporation into multicomponent subunit vaccines. In this study, we utilized the enhanced stability of bicontinuous nanospheres (BCN) to efficiently encapsulate MA for in vivo delivery to MA-specific T cells, both alone and in combination with an immunodominant Mtb protein antigen (Ag85B). Pulmonary administration of MA-loaded BCN (MA-BCN) elicited MA-specific T cell responses in humanized CD1 transgenic mice. Simultaneous delivery of MA and Ag85B within BCN activated both MA- and Ag85B-specific T cells. Notably, pulmonary vaccination with MA-Ag85B-BCN resulted in the persistence of MA, but not Ag85B, within alveolar macrophages in the lung. Vaccination of MA-BCN through intravenous or subcutaneous route, or with attenuated Mtb likewise reproduced MA persistence. Moreover, MA-specific T cells in MA-BCN-vaccinated mice differentiated into a T follicular helper-like phenotype. Overall, the BCN platform allows for the dual encapsulation and in vivo activation of lipid and protein antigen-specific T cells and leads to persistent lipid depots that could offer long-lasting immune responses.

Introduction

Vaccination is currently the most effective method for the prevention and eradication of infectious diseases. These efforts have been enhanced by the development of subunit vaccine formulations that include only immunodominant antigens from pathogens paired with select adjuvants. In contrast to attenuated and inactivated vaccines that contain diverse molecular components of pathogens, subunit vaccines are often preferred due to their simplicity, safety, stability, and scalability of production [1]. Yet the development of effective vaccines for several key endemic parasitic and bacterial pathogens has remained elusive. Antigen selection for subunit vaccines has thus far focused on protein molecules, overlooking a vast family of non-protein antigens such as lipids/glycolipids, metabolites, and phosphoantigens, which can be found on and within pathogens and can be recognized by various subsets of unconventional T cells [2].

Persistent antigen exposure is known to occur naturally after infection through antigen archiving by follicular dendritic cells (FDCs), which periodically present antigen to cognate B cells and aid in affinity maturation and somatic hypermutation [3, 4]. Antigen persistence can also support the maintenance of T cell response through the periodic priming of circulating T cells [5-10]. Subcutaneous vaccination has also been shown to lead to archiving of protein antigens within lymphatic endothelial cells (LECs), which induce the development of protective T cell immunity [11, 12]. However, little is known about how other vaccination routes and vaccine formulations may affect antigen archiving and persistence.

Lipid and glycolipid antigens can be presented by group 1 (CD1a, b, and c) and group 2 (CD1d) CD1 molecules to CD1-restricted T cells. Unlike MHC molecules, CD1 molecules exhibit limited polymorphism [13]. Therefore, CD1-restricted microbial lipid antigens are likely to be recognized by most individuals, making them attractive targets for vaccine development [14]. Group 1 CD1-restricted T cells have diverse T cell receptors and can recognize a variety of lipid and glycolipid antigens [15]. Microbial lipid-specific group 1 CD1-restricted T cells have been identified from several pathogens, including Mtb and other mycobacterial species: Staphylococcus aureus (SA), Borrelia burgdorferi, Salmonella typhimurium, and Brucella melitensis [15]. CD1b-restricted T cells specific to Mtb lipids including mycolic acid (MA), an immunodominant lipid antigen[16] and component of the mycobacterial cell wall, have been identified in patients with tuberculosis (TB) [17, 18]. As group 1 CD1 molecules are not expressed in mice, we have previously developed a human CD1 transgenic mouse model (hCD1Tg). [19] Using hCD1Tg mice, we have shown that transgenic CD1b-restricted DN1 T cells specific for MA could protect against Mtb infection [20], making these potential antigens of interest in a subunit vaccine against TB.

The primary goal of a vaccine is to induce the differentiation of the adaptive immune system such that upon re-exposure to the antigen, a faster and larger response occurs, halting pathogen spread. While conventional T cells and B cells are known to undergo this differentiation, unconventional T cells such as NKT cells do not possess a comparable memory phenotype. The presence of Mtb lipid-specific group 1 CD1-restricted memory T cells has been described in both human and mice [17-19]. However, the properties of memory group 1 CD1-restricted T cells, which share similar developmental pathway as NKT cells, are not known.

Lipid antigens are difficult to formulate and administer due to their hydrophobic nature. To assess the utility of MA as a component of a subunit vaccine, we previously encapsulated MA in poly(ethylene glycol)-block-poly(propylene sulfide) (PEG-b-PPS) micellar nanocarriers (MA-MC) [21]. Nanocarriers composed of PEG-b-PPS possess several unique benefits compared to other delivery platforms, including high stability in aqueous solution [22], minimal background immunostimulation [23], and triggered release of cargo when exposed to physiological levels of reactive oxygen species [24-26]. We showed MA-MC could effectively activate adoptively transferred DN1 T cells and elicit polyclonal group 1 CD1-restricted T cell response in hCD1Tg mice after intranasal delivery [21]. While effective for stable loading of lipophilic and amphiphilic payloads, micellar nanocarriers, do not allow for facile encapsulation of hydrophilic payloads like protein antigens.

Here, we build upon our prior lipid antigen vaccine formulation by employing a nanocarrier platform engineered for the dual loading of both lipid and protein antigens. Bicontinuous nanospheres (BCN) are polymeric analogs to lipid cubosomes and thus possess a highly organized internal cubic organization containing both intertwined hydrophobic bilayers for lipid loading as well as hydrophilic aqueous channels for protein loading [27]. Using the flash nanoprecipitation (FNP) technique [28, 29], we fabricated MA-loaded BCN (MA-BCN) and found them to have superior stimulatory capacity in vivo compared to MA loaded poly(D,L-lactide-co-glycolide) (PLGA) nanocarriers (MA-PLGA), which is a widely used nanocarrier platform for Mtb vaccine development [30, 31]. Furthermore, we found that after MA-BCN vaccination, MA could persist for 6 weeks post-vaccination within alveolar macrophages, a phenomenon which was dependent on the presence of an encapsulating vector, nanocarrier or bacterial, but not the route of vaccination. Due to the enhanced stability of the BCN architecture that can support the loading of diverse and multiple payloads [22, 32], we were able to efficiently co-encapsulate within BCN both MA and Ag85B, an immunodominant protein antigen of Mtb [33]. Interestingly, while both antigens activated their corresponding antigen-specific T cells, only MA resulted in antigen persistence, suggesting a potentially unique characteristic of subunit vaccines containing lipid antigens.

Results

BCN morphology preserved after MA loading

We scalably assembled spherical BCN with and without loaded MA using flash nanoprecipitation (FNP) as previously described [29]. Dynamic light scattering (DLS) showed that both MA-BCN and BCN were consistent in size (364 ±19 nm and 354 ± 14 nm, respectively), and were monodisperse based on their polydispersity indices (PDIs) of 0.24 ± 0.04 and 0.21 ±0.05, respectively (Fig 1A). We next verified that the BCN formulation maintained its characteristic interconnected aqueous channels using cryogenic transmission electronic microscopy (cryo-TEM) (Fig 1B) and negative staining transmission electron microscopy (TEM) (Fig 1C). Using small angle X-ray scattering (SAXS) studies, we confirmed the internal cubic organization of BCN. Bragg peaks at the √2, √4, and √6 ratios show that the primitive type of cubic internal organization was preserved between MA-BCNs and BCNs (Fig 1D). Thus, MA encapsulation did not disturb the BCN architecture. We also manufactured a PLGA nanocarrier formulation (PLGA-NP) with and without MA encapsulation. PLGA-NP morphology were confirmed using cryoTEM (Suppl Fig 1A) and SAXS (Suppl Fig 1B) and the size of blank PLGA-NP and MA-loaded PLGA-NP (MA-PLGA) were found to be 169 ± 4 and 163 ± 5 nm, respectively, with PDIs of 0.18 ± 0.04 and 0.19 ± 0.12, respectively (Fig 1E). Using the coumarin derivatization method [21], we found that BCNs could more efficiently encapsulate MA than PLGA-NPs (Fig 1F). We previously noted that the highly stable cubic architecture of BCN can lead to the retention of cargo within the endolysosomal compartment [22]. Indeed, Texas Red-Dextran loaded BCN were mostly found in lysosomes while Texas Red-Dextran loaded PLGA-NP could be found both within the lysosome and cytosol of cells (Supp Fig 2A, B).

Physicochemical characterization of BCN and PLGA-NP.

(A) Dynamic light scattering (DLS) analysis of blank BCN and MA-BCN. (B) CryoTEM of MA-BCN (scale = 500 nm). (C) Negative staining TEM images of MA-BCN. (D) SAXS of blank BCN and MA-BCN. (E) DLS analysis of blank PLGA and MA PLGA. (F) MA encapsulation efficiency for MA-BCN and MA PLGA. N = 3 per condition. Data represented as mean ± SD.

MA-BCN effectively activated MA-specific T cells in vitro and in vivo.

(A, B) BMDCs were pulsed for 18 hrs. with selected nanoparticles at various concentrations, co-cultured for 48 hrs. with DN1 T cells, and T cell activation was measured. (A) Percentage of CD69 and CD25-expressing DN1 T cells. (B) IFN-g production measured by ELISA. (N=3). (C, D) ELISPOT of IFN-g from MA-specific human M11 T cells cultured for 16 hours with monocyte-derived dendritic cells with selected nanoparticles at high (C) and low (D) concentrations. Data are representative of two independent experiments and displayed as mean ± SEM. Statistical analysis: 2-way ANOVA, significance designated for MA-BCN vs free MA. (E-G) hCD1Tg mice were IN vaccinated with selected nanoparticles and CellTrace Violet-labeled DN1 T cells were adoptively transferred the next day. After 1 week, DN1 T cell activation and proliferation was measured. (E) Representative FACS plots of DN1 T cells in the LN. Percentage of proliferating (F) and CD44-expressing (G) DN1 T cells in the LN and lung. Data represented as mean ± SEM. Statistical analysis: 2-way ANOVA. *p<0.05, **p<0.01, ***p<0.001, ****p< 0.0001.

MA-BCN activate MA-specific T cells in vitro and in vivo

As PLGA has served as a component of a wide range of vaccine formulations, including Mtb vaccines [30, 31], we benchmarked PEG-b-PPS BCN against PLGA-NP as a nanocarrier platform for lipid antigens. Both MA-BCN and MA-PLGA were highly effective in activating both mouse (DN1) and human (M11) MA-specific T cells in vitro and inducing IFN-ψ production (Fig 2A-D). In particular, MA-BCN were significantly better at activating MA-specific T cells compared to free MA at equivalent concentrations (Fig 2A, B and D). In comparison to MA-BCN, MA-PLGA was significantly more effective at stimulating mouse MA-specific T cells in vitro at lower concentrations. Interestingly, even in the absence of MA, PLGA-NP showed strong stimulation particularly for M11 T cells. These results demonstrate that MA encapsulation within BCN enhances its ability to stimulate antigen-specific T cells and reveal a considerable background, and thus difficult to control, stimulatory effect of blank PLGA-NP. In contrast, blank BCN showed little background immunostimulation [23], allowing a more controlled dose-dependent increase in T cell responses as the loaded MA concentration increased.

We next tested the ability of MA-loaded and unloaded BCN and PLGA to stimulate DN1 T cells in vivo. Intranasal (IN) vaccination of hCD1Tg mice was followed by the adoptive transfer of DN1 T cells. Surprisingly, we found that vaccination with MA-BCN induced a significantly higher percent of proliferation and activation in DN1 T cells than MA-PLGA in the draining lymph nodes (LN) and lungs at 1 week post-vaccination (Fig 2E-G). The extent of cell proliferation induced by MA-PLGA was not significantly different from that of blank PLGA (Fig 2F). Since MA-BCN effectively stimulated MA-specific T cells in vivo while MA-PLGA did not, we focused on further characterizing and assessing vaccination via solely MA-BCN formulations.

MA persists while Ag85B does not after vaccination with Ag85B-MA-BCN

Given the lack of knowledge regarding BCN and lipid antigen kinetics, it would be of interest to test whether delivered antigen could persist in the mouse weeks post-vaccination (Fig 3A). We switched from intranasal vaccination to intratracheal (IT) vaccination, as it allowed for the reliable delivery of a larger volume of the vaccine. We found that a significant proportion of the adoptively transferred DN1 T cells could proliferate and become activated in the LNs, lung, and spleen of MA-BCN-vaccinated mice 6 weeks post vaccination (Fig 3B-D). This activation and proliferation were MA-specific, as no activation and proliferation of Ag85B-specific p25 Tg (p25) T cells were observed[34].

Vaccination with MA-BCN leads to antigen persistence 6 weeks post-vaccination.

hCD1Tg mice were IT vaccinated with MA-BCN or BCN. 6 weeks later, CellTrace-labeled p25 and DN1 T cells were adoptively transferred, and T cell activation was measured after 1 week. (A) Experimental design. (B) Representative FACS plots of p25 and DN1 T cells in the LN. Percentage of proliferating (C) and CD44-expressing (D) p25 and DN1 T cells in the LN, lung, and spleen. N = 3 or 6 per condition. Outliers were identified through Grubbs method with alpha=0.05 with one sample removed. Data represented as mean ± SEM. Statistical analysis: 2-way ANOVA. *p < 0.05, **p < 0.01, ***p < 0.001, ****p < 0.0001.

Since BCN allow for the co-loading of lipid and protein antigens, we assessed the ability of dual loaded Ag85B-MA-BCN to activate antigen-specific T cells as well as lead to antigen persistence. We co-loaded Ag85B and MA in BCN and found Ag85B-MA-BCN to have a diameter of 380 nm and PDI of 0.223, comparable to MA-BCN (Fig 4A). Encapsulation efficiency was also relatively high at 70%.Vaccination of hCD1Tg mice with Ag85B-MA-BCN activated and induced proliferation of both p25 and DN1 T cells in the LN, lung, and spleen (Fig 4B-D) at one-week post-vaccination. In addition, DN1 T cells were still able to be activated 6 weeks post-vaccination in the context of an Ag85B-MA-BCN vaccination, while p25 T cells did not show increased activation or proliferation compared to blank BCN control at this time point (Fig 4E-H). Therefore, MA, but not Ag85B, appears to persist in the context of an Ag85B-MA-BCN vaccination.

Ag85B and MA dual loaded BCNs activate antigen-specific T cells without Ag85B persistence.

(A) DLS analysis of blank BCN and Ag85B-MA-BCN. (B-D) hCD1Tg mice were IT vaccinated with Ag85B-MA-BCN or BCN and CellTrace-labeled p25 and DN1 T cells were adoptively transferred the next day. After 1 week, T cell activation and proliferation was measured. (B) Representative FACS plots of p25 and DN1 T cells in the LN. Percentage of proliferating (C) and CD44-expressing (D) p25 and DN1 T cells in the LN, lung, and spleen (N = 4 or 5). (E-G) hCD1Tg mice were IT vaccinated with blank BCN or Ag85B-MA-BCN. 6 weeks later, CellTrace-labeled p25 and DN1 T cells were adoptively transferred, and T cell activation was measured after 1 week. (E) Representative FACS plots of p25 and DN1 T cells in the LN. (F) Percentage of proliferating p25 and DN1 T cells. (G) Percentage of CD44-expressing p25 and DN1 T cells in the LN and lung. (N = 4). Data represented as mean ± SEM. Statistical analysis: 2-way ANOVA. ns = not significant, *p<0.05, **p<0.01, ***p<0.001, ****p< 0.0001.

MA persistence was dependent on encapsulation but not route of delivery or delivery vector

To determine whether BCN encapsulation contributed to MA persistence, we designed an experiment that would allow direct comparison between MA-BCN and free MA. To account for their differential efficiency in stimulating DN1 T cells, we pulsed bone marrow-derived dendritic cells (BMDCs) from hCD1Tg mice with MA or MA-BCN at 10 μg/mL and 5 μg/mL MA concentration, respectively (Fig 5A). hCD1Tg mice were immunized with MA or MA-BCN-pulsed BMDCs at either 1 week or 6 weeks before the adoptive transfer of DN1 T cells, allowing comparison of short-term vs. long-term vaccination conditions. We found that while there were no significant differences in DN1 T cell activation and proliferation between mice vaccinated with free MA and MA-BCN pulsed BMDCs at 1-week post-vaccination, after 6 weeks, DN1 T cell response could only be detected in the MA-BCN vaccination condition (Fig 5B, C). Thus, BCN encapsulation contributed to the persistence of MA. We then tested whether the BCN structure could contribute to antigen persistence by comparing PEG-b-PPS BCNs to PEG-b-PPS MCs using a similar experimental set up (Suppl Fig 3). We found that MA-BCN and MA-MC overall had similar ability to activate DN1 T cells in both short-term and long-term vaccination time points (Suppl Fig 3B, C). Thus, the structure of the encapsulating NP did not play a significant role in antigen persistence of MA.

Route of vaccination or vector type do not affect antigen persistence.

hCD1Tg BMDCs were pulsed with MA-BCN or MA at concentrations of 5 μg/mL and 10 μg/mL, respectively. hCD1Tg mice were IT vaccinated with MA or MA-BCN pulsed BMDCs at 6 weeks or 1 week prior to the adoptive transfer of CellTrace-labeled DN1 T cells. T cell activation and proliferation was measured 1 week after adoptive transfer. (A) Experimental design. (B, C) Percentage of proliferating (B) and CD44-expressing (C) DN1 T cells in the LN, lung, and spleen of vaccinated mice. N = 4 per condition. N = 4 per condition. hCD1Tg mice were vaccinated SC (subcutaneously) or IV (intravenously) with blank BCN, MA-BCN, or attenuated Mtb strain and 6 weeks later CellTrace-labeled DN1 T cells were adoptively transferred. (D) Representative FACS plots of DN1 T cells in the LN of mice vaccinated with indicated conditions. (E) Percentage of proliferating DN1 T cells and (F) Percentage of CD44-expressing DN1 T cells in the LN and spleen of mice vaccinated with Blank-BCN (SC), MA-BCN (IV and SC), and attenuated Mtb. N = 3-5 per condition. Data represented as mean ± SEM. Statistical analysis: 2-way ANOVA. ns = not significant, *p<0.05, **p<0.01, ***p<0.001, ****p<0.0001.

To investigate if the IT route of vaccination played a role in the development of antigen persistence, we tested intravenous (IV) and subcutaneous (SC) vaccination routes with MA-BCN. DN1 T cell activation and proliferation were detected at 6-weeks post-vaccination through both routes of vaccination (Fig 5D-F), suggesting that the antigen persistence is not unique to IT vaccination. To assess whether Mtb lipid antigen persistence could also be observed in mice vaccinated with an attenuated strain of Mtb, we vaccinated hCD1Tg mice SC with Mtb mc26206 strain and tested for the persistence of both Ag85B and MA after 6 weeks. The Mtb mc26206 strain displays less virulence in mouse studies than BCG but still confers long-term protection against virulent Mtb challenge equivalent to BCG[35]. We confirmed that no bacterial burden remained at the 6 weeks time point (Suppl Fig 4A). Similar to MA-BCN-vaccinated mice, DN1 T cells were activated and proliferated in the draining lymph nodes and spleen of mice vaccinated with the attenuated Mtb strain (Fig 5D-F). In line with our findings with BCN, no proliferation was seen in p25 T cells (Suppl Fig 4B-D). These results suggest that incorporation of MA in a nanocarrier vaccine allows for the mimicry of the lipid antigen persistence induced by vaccination with an attenuated Mtb vaccine.

MA persists within alveolar macrophages in the lung

To identify the location of antigen persistence, we determined the biodistribution of BCN after IT vaccination using BCN loaded with the hydrophobic dye DiD (DiD-BCN). Since DiD is a lipophilic dye which, like MA, stably loads into the BCN bilayers, we could track the localization of BCN without having to perform additional chemical modifications. We found that IT-administered DiD-BCN were located almost exclusively within the lung at all time points tested (4 h, 24 h, 48 h, 6 days, and 6 weeks) (Fig 6A). We were able to detect fluorescence through both in vivo imaging system (IVIS) and flow cytometry at 6-weeks post-vaccination that was significantly above unvaccinated control (Fig 6A, B), despite DiD quickly losing fluorescence upon mixture into aqueous environments [36], suggesting that some DiD-BCN structure remained intact at this time point. Flow cytometric analysis (Suppl Fig 5) revealed that DiD BCN was primarily found within alveolar macrophages (AMs) at both early and late time points with neutrophils and DCs showing some fluorescence at early time points (4 and 24 h) (Fig 6C).

Persistent MA remains in part due to encapsulation inside alveolar macrophages and activates T cells through DCs.

B6 or hCD1Tg mice were IT vaccinated with BCN loaded with a hydrophobic fluorescent dye (DiD) at 6 weeks, 6 days, 48 hours, 24 hours, or 4 hours prior to the experiment. LN, lung, and spleen were analyzed using (A) In Vivo Imaging System (IVIS). Single cell suspension was then obtained and presence of DiD BCN was quantified in (B) total lung, and (C) CD45- cells, neutrophils (Ly6G+), alveolar macrophages (CD11c+SiglecF+), DCs (CD11c+), monocytes (CD11b+CD11c-), B cells (CD19+), T cells (CD3+), NK cells (NK1.1+), eosinophils (CD11c-SiglecF+) by flow cytometry. Gating strategy exemplified in Suppl Fig 4. hCD1Tg mice were IT vaccinated with either BCN or MA-BCN. After 6 weeks, alveolar macrophages were enriched from lungs using anti-SiglecF. Enriched or flow through cells were co-cultured with DN1 T cells in the presence or absence of hCD1Tg-expressing BMDCs for 48 hours, and DN1 T cells activation was measured. (D) Experimental design. (E) Percentage of CD25-expressing DN1 T cells. Data represented as mean ± SEM. Statistical analysis: 2-way ANOVA. ns = not significant, *p<0.05, **p<0.01, ***p<0.001.

We next tested whether MA likewise persisted primarily within AMs. 6-weeks after vaccination, we enriched for AMs using a column-based magnetic cell isolation system containing SiglecF antibody, which is highly expressed on AMs and a small population of eosinophils in the lung (Fig 6D). We co-cultured enriched AMs or the flow through with or without hCD1Tg-expressing BMDCs and DN1 T cells (Suppl Fig 6). While both enriched AM and flow through fractions from MA-BCN vaccinated mice co-culture with BMDCs could activate DN1 T cells (Fig 6E), co-culture with enriched AMs led to significantly higher DN1 T cell activation, suggesting AMs are the primary location of MA persistence. The ability of flow through to likewise lead to DN1 T cell activation suggests MA may also persist in other cell types or could be attributed to residual AMs in flow through. Furthermore, AMs could not themselves present the MA to DN1 T cells, noted by the lack of T cell activation in the absence of BMDCs, suggesting that MA may be transferred from AMs to DCs for presentation. In fact, CD1b is not expressed on AMs [19, 37] and therefore the presence of CD1b-expressing DCs may be necessary both in vitro and in vivo for T cell activation to occur. These data indicate that MA-BCN mainly persists within AMs in the lung after IT vaccination and CD1b-expressing DCs is required for antigen presentation and activation of MA-specific T cells.

Vaccination leads to DN1 T cell differentiating into T follicular helper-like T cells

To study the memory phenotype of DN1 T cells after MA-BCN vaccination, we constructed a mixed DN1 bone marrow (BM) chimera mouse model (DN1-hCD1Tg), since adoptively transferred DN1 T cells in unvaccinated mice cannot survive long term (Fig 7A). After vaccination, the percent of CD44+ DN1 T cells in the blood increased quickly and remained high until the last time point of 40 days (Fig 7B). Various classical memory T cell subsets have been characterized 1 month after initial antigen-encounter, we therefore determined the DN1 T cell phenotype 6-weeks post vaccination. We found that the most prevalent memory population within DN1 T cells were CD44+CD62L+, markers used to define central memory T cells, particularly in the LN (Fig 7C and Supp Fig 7A). To characterize the memory DN1 T cells, we performed RNA-seq analysis on sorted CD44+CD62L+ (memory) and CD44-CD62L+ (naïve) DN1 T cells from LNs of MA-BCN vaccinated DN1-hCD1Tg BM chimeric mice. We found that memory and naïve DN1 T cells clustered separately after principal component analysis (PCA), despite these samples coming from the same animals (Suppl Fig 7B). A total of 995 differentially expressed genes (DEGs) were identified of which 542 upregulated and 453 downregulated in the memory subset (Suppl Fig 7C). Next, we determined which T cell population memory DN1 T cells most resembled. Toward this end, we obtained data from ImmGen database (including terminally differentiated effector (Te), memory precursor (Tmp), central memory (Tcm), effector memory (Tem), regulatory (Treg))42, and two additional publications (follicular helper (Tfh)43, exhausted (Texh)44) and compared the respective DEG lists. Using PCA, we found that memory DN1 T cells clustered most closely to Tfh, Treg, and Texh (Fig 7D) and when this subset is isolated, most closely to Tfh cells (Fig 7E).

MA-BCN vaccination leads to DN1 T cell differentiation into T follicular helper-like T cells.

Mixed bone marrow chimeric mice were created by adoptive transfer of DN1 bone marrow and congenic CD45.1 hCD1Tg bone marrow to irradiated hCD1Tg mice. After 5 weeks, mice were vaccinated with either MA-BCN or PBS, and DN1 surface marker expression was monitored in the blood and in organs at 6 wks. (A) Experimental design. (B) Percent CD44+ DN1 T cells in the blood at various time points post vaccination. (C) Percent CD44+CD62L+ DN1 T cells in indicated organs at 6 weeks post vaccination. Memory (CD44+CD62L+) and naïve (CD44-CD62L+) DN1 T cells were sorted from MA-BCN vaccinated hCD1Tg-DN1 BM chimeras at 6 weeks post vaccination and subjected to RNAseq analysis. N=3 per condition (D, E) PCA of log fold change of gene expression for each T cell subset was performed relative to internal naïve T cell control. (F) Relative expression of key TFH cell and DEGs in memory subset. (G) Gene enrichment analysis of upregulated DEGs was performed using Metascape. (H) Representative FACS plots of DN1 T cells in the LN of mice. Percentage of (I) CXCR5+PD1+ or (J) KI67+ DN1 T cells DN1 T cells in LN, lung, and spleen. N = 3-5 per condition. Data represented as mean ± SEM. Statistical analysis: 2-way ANOVA. *p<0.05, **p<0.01, ****p < 0.0001.

Within the DEGs, we noted upregulation of key Tfh cell transcription factors BCL6 and TCF1 (TCF7), chemokines CXCR5 and CXCR4 and surface receptor ICOS, important for Tfh cell migration into the germinal center, and adaptor protein SAP (SH2D1A), important for T cell-dependent B cell immunity (Fig 7F)45. Inhibitory receptors PD-1 (PDCD1), TIGIT, CTLA4, and transcription factor EOMES, which are known to be expressed on Tfh cells were also upregulated. In addition, we observed downregulation of transcription factor KLF2, which can inhibit Tfh cell differentiation, and CCR7 and IL-7R, receptors known to have decreased expression in TFH cells. Consistent with our transcriptome analysis, we found increased percentage of CXCR5+PD-1+ DN1 T cells in MA-BCN vaccinated mice 6 weeks post-immunization (Fig 7G, H). Both our transcriptomic and experimental data highlighted that memory, but not naïve DN1 T cells, were proliferating, given the enrichment for the mitotic cell cycle process (Fig 7I) and increased KI67+ percentage within this population (Fig 7J). MA-BCN vaccination thus led to the differentiation of DN1 T cells into a proliferating Tfh-like T cell population.

Discussion

Research on the development of subunit vaccines containing lipid antigens is limited. This is in part due to a lack of research into appropriate lipid vaccine formulations and their associated properties. Some early evidence suggests that lipid vaccines could contribute to protection in Mtb challenge in guinea pig model [38, 39], which naturally express group 1 CD1 molecules [40]. Vaccination with a formulation composed of total Mtb lipid encapsulated within liposomes with either dimethyl-dioctadecyl-ammonium (DDA) and/or QS-21 adjuvants showed lower bacterial burden and pathology in lung and spleen compared to unvaccinated controls [38]. In another study, two Mtb lipids (diacylated sulfoglycolipids and the phosphatidylinositol dimannosides) encapsulated into liposomes formulated with DDA and trehalose 6,6’-dibehenateled (TDM) to a decreased bacterial burden in the spleen and overall improved pathology in lung and spleen [39]. This work motivated our current study, where we used hCD1Tg mice that express human group 1 CD1 molecules, including CD1b and MA-specific (DN1), as well as p25-specific T cells to investigate the use of nanocarriers coloaded with both Mtb lipid and protein antigens during vaccination (Fig 8).

Summary schematic illustrates MA antigen persistence and memory differentiation of antigen-specific T cells induced by vaccination with MA-loaded nanoparticle vaccine.

Mycobacterium tuberculosis (Mtb) derived mycolic acid (MA) lipid and Ag85B protein are encapsulated into bicontinuous nanosphere (BCNs) composed of PEG-b-PPS polymer. These MA-BCN or Ag85B-MA-BCN are used for intratracheal (IT) vaccination of human CD1 transgenic (hCD1Tg) mice. Following vaccination, dendritic cells (DCs) present the antigens, leading to the activation of CD1b-restricted MA-specific or MHCII-restricted Ag85B-specific T cells (DN1 or p25-specific T cells, respectively). Notably, MA can persist within alveolar macrophages and continue to activate MA-specific T cells even at 6 weeks post-vaccination. These MA-specific T cells have the capacity to differentiate into memory T cells with T follicular-like phenotype.

We found that PEG-b-PPS BCN could effectively encapsulate MA as well as the protein antigen Ag85B to elicit activation of antigen-specific T cells in vaccinated mice (Fig 4). To our knowledge, this is the first time a dual protein and lipid loaded nanocarrier has been shown to effectively stimulate both lipid and peptide antigen-specific T cells simultaneously in vivo. PEG-b-PPS BCN can encapsulate both hydrophobic and hydrophilic molecules, making it an excellent platform for subunit vaccine delivery as both antigens and adjuvants can be co-loaded and delivered to the same cellular targets. Prior work on TB vaccines has shown that the inclusion of several protein antigens spanning the Mtb life cycle along with adjuvants with multiple boosts is paramount in the development of a vaccine that can provide effective protection against Mtb [30, 41]. As we have established that PEG-b-PPS BCN can easily incorporate a variety of payloads, future work will involve determining the optimal antigen and adjuvant combination for providing protection in a Mtb challenge experiment.

As several formulations of PLGA-NP have been approved for clinical uses [42], we compared MA encapsulated by either BCN or PLGA-NP and found that while PLGA appeared to offer greater, yet less controllable stimulation in vitro (Fig 2), BCN exhibited significantly greater efficacy in vivo (Fig 2). The result is not completely unexpected as a more fragile PLGA-NP could allow for fast intracellular release and antigen presentation in vitro while also leading to significant antigen loss prior to APC interaction in vivo (Fig 3) [43]. Previous studies have documented PLGA’s innate pro-inflammatory effects, while other studies have highlighted the absence and suppression of inflammatory response [44-46]. As PLGA has not been previously used for lipid antigen delivery via IT vaccination, our findings suggest that PLGA may have limited effectiveness in such a vaccine system, although alternative formulations or polymer compositions could potentially yield improved results. In contrast and as previously observed for PEG-b-PPS nanocarriers [23], we showed that unloaded BCN have little inflammatory effect (Fig 2), allowing their immunomodulation to be dictated by the properties of the loaded antigen and adjuvant. Thus, PEG-b-PPS is a promising polymer for the formulation of subunit vaccines.

We proceeded with a pulmonary (IN or IT) vaccination approach in this study based on emerging evidence that mucosal vaccination can induce a stronger immune response in the lung against Mtb in humans [47] and improve protection following vaccination with nanocarriers in mice [48]. We found that in the pulmonary route of vaccination, BCN primarily were taken up and remained within alveolar macrophages in the lung (Fig 6). In fact, the ability of nanocarriers to target alveolar macrophages or other APCs may contribute to their distinct capacity for inducing antigen persistence, which was not seen with cellular carriers such as MA-pulsed BMDCs (Fig 5). Alveolar macrophages are the first line of defense of the lungs, responsible for keeping alveoli sterile by taking up inhaled particles and pathogens through phagocytosis [49-51]. Targeting these cells may be particularly beneficial in a TB vaccine design since they may also be the main site of Mtb latency [52]. While MA remained within alveolar macrophages in the lung, T cell activation appeared to occur prominently within the draining lymph nodes (Fig 2), suggesting that either MA was transported from the lungs to lymph nodes where the presentation took place or T cells activated in the lung migrated to the lymph nodes. The first option seems most likely, as CD1b-expressing DCs are more prevalent in the lymph nodes [53]. Furthermore, in Mtb infection models, it has been previously shown that apoptotic macrophage and vesicles could deliver mycobacterial antigens to DCs that then activate cognate T cells [54, 55]. Similarly, we showed that alveolar macrophages from MA-BCN vaccinated mice could only lead to T cell activation in the presence of BMDCs from hCD1Tg mice (Fig 6).

We demonstrated MA persistence in the lung 6 weeks post-vaccination using an in vivo antigen presentation assay with adoptive transfer of naïve MA-specific T cells. Sustained release of antigen through vaccination is thought to be key in the development of a robust adaptive immune response [56, 57]. Recently, spike protein was found within the draining lymph nodes of vaccinated human subjects 60 days after COVID-19 vaccination with mRNA-1273 or BNT162b2 [58]. In mice, pulmonary vaccination with adenovirus expressing influenza nucleoprotein supported long term maintenance of CD8 TRM (resident memory) cells [59]. Unlike in this work, antigen persistence did not occur after subcutaneous vaccination, suggesting a distinct mechanism of antigen archiving. In a different vaccine model, subcutaneous administration of ovalbumin paired with adjuvants polyl:C and anti-CD40 led to antigen archiving by LECs [11, 12]. Similar to our findings, antigen-specific T cell activation required cross-presentation by DCs [11, 60]. Both LECs and FDCs retain antigens within a non-degradative compartment [11, 61], suggesting that residual antigens in our model may similarly remain in the endosome within undigested BCN. We have previously demonstrated the enhanced stability of BCN within cell endosomes in vitro for extensive periods of time [22]. This possibility is corroborated by the finding that DiD, which is easily quenched in aqueous solution [36], remained fluorescent 6 weeks post-vaccination (Fig 6). The lack of persistence for Ag85B could be due to its leakage out from the aqueous pores of BCN, which we have observed for other proteins [29]. Peptide antigens may be modified to include a hydrophobic tail for stable retention within BCN and therefore allow for peptide antigen persistence as well. MA’s extreme hydrophobic surface, which correlates with its low permeability as a part of Mtb cell wall, may also contribute to its prolonged persistence in vivo [62].

After infection and vaccination, peptide antigen persistence has been shown to occur through archiving by binding to CD21 receptors on follicular dendritic cells (FDCs) in lymph nodes [3, 4]. MA antigen persistence may likewise be a natural phenomenon that can occur after infections, although its detection may be difficult given the lack of available assays to detect very small amounts of lipid antigen. T cells may, in fact, be more sensitive than any commercially available tool. While we determined that MA persists within AMs following pulmonary vaccination, this localization is likely dependent on the vaccination route. In the case of SC and IV routes, significant exposure of AMs to MA-BCN is unlikely to occur. Consequently, the exact localization of MA after vaccination via SC, IV, or attenuated Mtb remains to be determined, and it is plausible that other types of macrophages may be involved in this process. Additionally, we found that different forms of encapsulation (such as BCN, MC, attenuate Mtb) can induce antigen persistence, DC internalization alone was not sufficient to replicate this phenomenon (Fig 5). Thus, the question still remains regarding the necessary properties of the delivery vector for the persistence of antigens to occur. Overall, MA antigen persistence is a robust phenomenon which occurs independently of the vector and vaccine route of administration, and thus may play a role in the efficacy of a variety of vaccine strategies. Additional work will focus on whether lipid antigen persistence is found across the lipid family or specific to lipids with certain properties.

Antigen persistence can also occur in chronic infections in which it is known to induce T cell exhaustion [63]. Despite differentiating in an environment of persistent antigen exposure, memory DN1 T cells exhibited a gene expression profile more similar to that of Tfh cells rather than exhausted T cells (Fig 7). In particular, memory DN1 T cells upregulated the expression of TCF7 and BCL6 which are known to sustain a non-exhausted phenotype in CD8 T cells during chronic viral infections [64, 65]. Previous studies on antigen archiving have suggested that the mere persistence of antigen is insufficient for exhaustion to develop, and the inflammatory context likely plays a role in T cell differentiation [11]. In many subunit vaccines, virulence factors are used as antigens to target and suppress the pathogen’s capacity to infect the host. The persistence of lipid virulence factors could raise concerns regarding the safety of lipid vaccines. However, it is worth noting that otherwise toxic compounds such as the lipid TDM have been successfully incorporated into numerous vaccines without any known safety concerns to date [66-71]. The absence of reported safety issues may be attributed to limited research or the administration of low doses of TDM. It is important to highlight that while MA play a role in immune evasion, MA without additional functional groups has not been shown to elicit similar toxic effects as TDM [62].

The use of memory T cells in the setting of antigen persistence may be debatable, given that memory response has historically referred to the adaptive immune response in the setting of antigen clearance [72]. However, given that the persistence of MA occurs even in the setting of attenuated Mtb vaccination, Mtb lipid-specific long-term adaptive immunity will likely naturally exist in an antigen persistent environment. We found that DN1 T cells in this setting expressed key markers of Tfh cells such as CXCR5, PD-1, BCL6, and TCF1 (Fig 7). T cells differentiate into Tfh cells through the expression of master transcription factor regulator BCL6, which directs the upregulation of chemokine receptors CXCR5 and CXCR4 and thus migration into the germinal center[73, 74]. In the germinal center, Tfh cells provide help to B cells for affinity maturation and class switching. Other unconventional T cells are known to behave similarly such as CD1d-restricted NKT cells, which can acquire a Tfh phenotype and provide B cell help through both cognate and non-cognate interactions[75]. Additional work will look at whether DN1 or other group 1 CD1-restricted T cells can functionally act as Tfh cells to provide B cell help, as may be suggested by the existence of IgG antibodies against MA and glycolipid lipoarabinomannan in TB patients [76-78].

While persistence of protein antigen has been shown to improve the quality of protective immunity [7, 11, 79, 80], it is not yet known how lipid antigen persistence may affect the protective efficacy of lipid-specific T cells. At the very least, antigen persistence increases the probability of T cell/DC encounters by increasing the time that the antigen is available. Additional work will need to address the function of lipid-specific T cells differentiated in antigen persistent environments. As the ultimate goal of this work is to improve the efficacy of existing subunit vaccines, future work will assess the protective capacity of lipid antigen vaccines paired with additional protein antigens and adjuvants.

Materials and Methods

Ethics statement

This study was carried out in accordance with the recommendations in the Guide for the Care and Use of Laboratory Animals of the National Institutes of Health. The protocol was approved by the Institutional Animal Care and Use Committee of Northwestern University (Protocol number: IS000011717).

Poly (ethylene glycol)-block-poly (propylene sulfide) polymer synthesis

Poly (ethylene glycol)17-block-poly-(propylene sulfide)75 (PEG17-b-PPS75) copolymer was synthesized as previously reported [81, 82]. Briefly, monomethoxy PEG (molecular weight=750 Da) was functionalized with mesyl chloride and substituted with thioacetate to produce a PEG acetate initiator for the polymerization for PPS.

Bicontinuous nanosphere fabrication via flash nano precipitation

BCN were produced using FNP technique as previously described[81]. In brief, PEG17-b-PPS75 BCN polymer and any hydrophobic cargo was dissolved in THF and loaded onto a syringe. The hydrophilic cargo was loaded onto a separate syringe, and both impinged against one another through the confined impingement mixer into scintillation vial. The vial was placed in a vacuum desiccator overnight to remove the organic solvent. NPs were filtered through LH-20 column with PBS as an eluent and suspended in PBS for later use. A natural mixture of Mtb MA consists of α-, keto-, and methoxy forms of MA with chain lengths of 72-86 carbon atom was obtained from Sigma (M4537, Sigma-Aldrich, St. Louis, MO) [83]. MA encapsulation efficiency was analyzed using coumarin derivatization as previously described [21]. MA was loaded at a concentration of 100 µg/mL of total BCN formulation and DiD was loaded at 5 µg/mL total BCN formulation. MA loading capacity in BCNs was 2%. To measure the encapsulation efficiency of Ag85B, Ag85B-MA-BCN were centrifuged at 10,000 g for 10 minutes. The supernatant was collected and concentrated using Amicon® Ultra-4 10K centrifugal filter devices. To determine the unencapsulated protein concentration, the supernatant was incubated with Pierce™ BCA Protein assay reagent following the manufacturer protocol and the absorbance was measured at 660 nm using a SpectraMax M3 multi-mode microplate reader (Molecular Devices, LLC). The percentage encapsulation efficiency was calculated as a percent of encapsulated protein to the total amount of protein added. Ag85B was encapsulated at a concentration of 140 µg/mL and loading capacity was found to be 2.88%.

Poly (lactic-co-glycolic acid nanoparticle (NP) fabrication

PLGA-NP were prepared by the Oil-in-Water single emulsion method. Briefly, PLGA (20 mg in 800 µL dichloromethane) organic phase was emulsified using an ultrasonic processor with aqueous phase containing 2 mL of polyvinyl alcohol (PVA) solution (2.5% w/v) to form an emulsion. The emulsion was then added dropwise into 4 mL of stirring 0.5% w/v PVA solution at room temperature to evaporate the organic solvent. NPs were collected after 4 hours of stirring followed by centrifugation at 10,000 x g for 10 minutes. After centrifugation, NPs were washed twice with cold water to remove residual PVA and redispersed in PBS. Hydrophilic (Texas red-Dextran) or hydrophobic (MA) loading was performed by adding the payloads to the aqueous and organic phases, respectively.

Nanocarrier characterization

The nanocarrier size (z-average diameter) and polydispersity was measured using dynamic light scattering (DLS) using a Nano 300 ZS Zetasizer (Malvern Panalytical, Malvern, U.K.). BCN and PLGA morphology was visualized with cryogenic transmission electron microscopy (cryoTEM) as previously detailed [21]. Transmission electron microscopy (TEM) was performed to image BCN using uranyl acetate as a negative stain as previously reported [22]. BCN aggregate structure and internal morphology was characterized with small angle X-ray scattering (SAXS). These studies were performed at the DuPont-Northwestern-Dow Collaborative Access Team beamline at Argonne National Laboratory’s Advanced Photon Source with 10 keV (wavelength λ = 1.24 Å) collimated X-rays, as described previously[81].

Confocal imaging

Confocal images of RAW 264.7 macrophages stained with lysosomal dye Lysotracker™ (green) and NucBlue™ stain (blue) following incubation with Texas Red-labeled PLGA-NP and BCN (red) for 8 hr. The cells were then imaged within a humidified chamber using a 63X oil-immersion objective on a SP5 Leica Confocal Microscope using HyD detectors and lasers. Data analysis was performed using ImageJ software.

Mouse strains

Human CD1 transgenic mice (hCD1Tg) in C57BL/6 background or MHC II-deficient background [19] and CD1b-restricted MA-specific TCR transgenic mice on Rag-/- background (DN1Tg/hCD1Tg/Rag-/-) [20] were generated in our lab as previously described. P25-specific TCR transgenic mice (Jackson lab) were crossed onto Rag-/- background. Both males and females were used.

Antibodies and flow cytometry

For cell surface staining, cells were pre-incubated with 2.4G2 Fcγ RII/RIII blocking mAb for 15 minutes and then stained with the appropriate combinations of mAbs diluted in HBSS + 2% FBS for 30 minutes at 4°C. Cells were analyzed on a BD FACS CantoII, or Cytek Aurora Spectral Cytometer and data were processed with FlowJo software (TreeStar). The complete list of antibodies use is as follows: anti-human Vβ5.1 (LC4, Invitrogen), anti-TCRβ (H57-597, BioLegend), anti-CD44 (IM7, BioLegend), anti-Vβ11 (KT11, BioLegend), anti-CD45.2 (104, BioLegend), anti-CD45.1 (A20, BioLegend), anti-SiglecF (E50-2440, BD Bioscience), anti-CD45 (30-F11, BioLegend), anti-CD11c (N418, BioLegend), anti-CD11b (M1/70, BioLegend), anti-Ly6G (1A8, BioLegend), anti-CD19 (6D5, BioLegend), anti-NK1.1 (PK136, BioLegend), anti-CD3 (17A2, BioLegend), Zombie Red (BioLegend), anti-mouse CD185/CXCR5 (L138D7, Biolegend), anti-mouse CD279/PD-1 (29F.1A12, Biolegend), and anti-Ki67 (SolA15, Invitrogen).

In vitro nanoparticle and lipid titration with MA-specific DN1 T cells

Nanoparticle and lipid titrations were performed as previously described [21]. In brief, hCD1Tg MHC II-/- bone marrow-derived dendritic cells (BMDCs) were pulsed with varying quantities of nanoparticle of sonicated MA and then co-cultured for 48 hours with DN1 T cells isolated from lymph nodes (axial, brachial, submandibular, inguinal) of DN1Tg/hCD1Tg/Rag-/- mice. The supernatant was used for IFN-ψ enzyme-linked immunosorbent assay (ELISA) performed as described previously [21] and cells were stained for activation.

In vitro nanoparticle and lipid titration with human cells via ELISPOT assay

Hydrophobic polyvinylidene fluoride 96-well plates (Millipore, Bedford, MA) were coated with anti-human IFN-γ mAb 1-D1K (Mabtech, Cincinnati, OH) diluted 1:400 in PBS (Gibco, Waltham, MA) and incubated at 4°C overnight. The following day, mAb was removed and each plate was washed and incubated at room temperature for 2 hours in the presence of media. Human monocyte-derived dendritic cells and MA-specific M11 T cells were seeded 50,000 cells and 2,000 cells per well, respectively, together with serially diluted MA and NPs. After 16-hour incubation, plates were washed and incubated with biotinylated anti-human IFN-g (7-B6-1), diluted 1:3000 in PBS + 0.5% FBS, for 4 hours. Plates were then washed with PBS 5 times and incubated for 2 hours with ExtrAvidin-Alkaline phosphatase diluted 1:1000 in PBS + 0.5% FBS. Spots were visualized after incubation with the BCIP/NBT substrate for up to 20 minutes and then imaged with ImmunoSpot reader v.2.0.

Mouse vaccination

For both intranasal (IN) and intratracheal (IT) vaccination, mice were anesthetized using inhaled isoflurane. For IN vaccination, 25 µL of liquid was administered at the nostrils of the mice for a total of 1 μg of MA per mouse. For IT vaccination, mice were administered 25 or 50 μL of liquid for a total of 2.5 μg of MA with or without 2.0 μg of Ag85B. For BMDC vaccinations, hCD1Tg BMDCs were cultured in GM-CSF and IL-4 for 5 days[84] and pulsed with MA-BCN (5 μg/mL of MA) or free MA (10 μg/mL of MA) for 24 hours. Cells were washed and resuspended in PBS and IT vaccinated at a dose of 1x106 BMDCs/mouse. Intravenous vaccination was administered by tail vein injection. Subcutaneous vaccination was administered between the shoulders over the neck portion of the mouse.

Vaccination with attenuated Mtb

Attenuated Mtb strain (H37Rv τ1panCD τ1leuCD referred to as mc2 6206) was kindly provided by Jacobs’ Laboratory at Albert Einstein College of Medicine. Bacteria were grown as previously described in supplemented Middlebrook 7H9 media [35, 85]. Mice were vaccinated subcutaneously with 1x106 CFU in 100 μL of PBS. Mycobacterial burden quantification was performed by plating serial dilutions of lung or spleen homogenate on Middlebrook 7H11 agar plates [35, 85].

T cell Adoptive transfer and cell preparation

DN1 or p25 T cells were isolated from lymph nodes of DN1Tg/hCD1Tg/Rag-/- or P25Tg/Rag-/- mice, respectively. Cells were labeled with CellTrace™ Violet reagent (Invitrogen™) as per manufacturer’s instructions and 3-5x106 cells were injected into hCD1Tg mice either 1 day before vaccination (short term) or 6 weeks post-vaccination (long term). On day 7 post-adoptive transfer, draining lymph nodes (axial, brachial, mediastinal), lung, and spleen were obtained. Single-cell suspensions were prepared by mechanical disruption in HBSS/2% FBS. The lung was digested with collagenase IV (1mg/ml; Sigma-Aldrich) and DNase I (30µg/ml; Sigma-Aldrich) for 30 minutes at 37°C before the disruption.

Biodistribution imaging

BCN were loaded with DiD dye (Thermo Fisher) and administered to mice IT at 4 hours, 24 hours, 48 hours, 6 days, and 6 weeks prior to the final time point, at which all mice were sacrificed and analyzed together. Lung, spleen, and draining lymph nodes were harvested and imaged using a near-IR in vivo Imaging System (IVIS; Center for Advanced Molecular Imaging, Northwestern University). The single-cell suspension was then prepared from indicated organs and cells were stained for flow cytometric analysis.

Alveolar macrophage enrichment and T cell co-culture

hCD1Tg mice were immunized IT with either BCN or MA-BCN (2.5 μg of MA/mouse). 6 weeks later, lungs were isolated and single-cell suspension was stained with anti-SiglecF-PE (E50-2440, BD Bioscience). SiglecF-positive cells were enriched using anti-PE MultiSort Kit (Miltenyi Biotec). 1x105 flow through and AM enriched fractions were co-cultured with 1x105 hCD1Tg BMDCs and 3x105 DN1 T cells for 48 hours and cells were stained for the expression of activation markers.

Generation and vaccination of BM chimeras

Bone marrow cells from DN1Tg/hCD1Tg/Rag-/- (CD45.2) and hCD1Tg mice (CD45.1) were depleted of mature T cells using anti-Thy-1.2 (AT83.A-6) plus rabbit complement (Cedarlane Laboratories). Equal numbers of cells (5x106) were adoptively transferred into hCD1Tg (CD45.1) mice irradiated with 900 rads. After 6 weeks, BM chimeric mice were intratracheally vaccinated with either 2.5 μg of MA in MA-BCN or equivalent volume of PBS and the activation status of DN1 T cells were monitored in the blood 1-2 times per week until experiment completion.

RNA-Seq analysis

CD44+CD62L+ and CD44-CD62L+ DN1 T cells were sorted from LNs of DN1-hCD1Tg BM chimera mice vaccinated with MA-BCN at 6 weeks post vaccination by BD FACS Aria with purity >98%. RNA was extracted with RNAeasy® mini kit (Qiagen). Libraries were generated by using the Illumina Truseq preparation kit and sequenced on HiSeq4000. Reads were analyzed with Ceto pipeline (https://github.com/ebartom/NGSbartom) using STAR and HTseq for alignment on mm10 mouse genome and reading counting, as described previously [86]. Paired differential expression analysis was performed using DESeq2 to account for original mouse sourcing[87]. All downstream analysis was performed in R and figured generated using ggplot. Gene enrichment analysis was performed using Metascape [88]. Comparisons to other T cell profiles were performed using raw transcriptome data from ImmGen database[89], and two additional publications for follicular helper (TFH)[90], exhausted (Texh)[91]. Analysis was performed as described above to obtain gene expression fold change values relative to internal naïve condition.

Data accessibility

The data discussed in this publication have been deposited in NCBI’s Gene Expression Omnibus and are accessible through GEO Series accession number GSE226075 (https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE226075).

Statistical analysis

Statistical analyses were performed using Prism software 9 (GraphPad software, Inc.). Multi-group comparisons were done using two-way analysis of variance (ANOVA) with significance determined by Tukey’s multiple comparison. Where appropriate, outliers were identified through Grubbs method with alpha=0.05. Statistically significance denoted as ns = not significant, *p<0.05, **p<0.01, ***p<0.001, and ****p< 0.0001.